---
title: "PRSice-2 workflow"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(data.table)
library(pROC)
```

# Overview

This notebook prepares GWAS summary statistics and phenotype/covariate files, runs PRSice-2, and evaluates the resulting polygenic risk score (PRS) for a binary or continuous phenotype.

The workflow assumes that the target genotype data are available in PLINK binary format (`.bed`, `.bim`, and `.fam`) and that the GWAS summary statistics and phenotype data have already undergone appropriate quality control.

# Required software and input files

## PRSice-2

Download PRSice-2 from the official repository:

- Repository: <https://github.com/choishingwan/PRSice>
- Documentation and downloads: <https://choishingwan.github.io/PRSice/>

For Windows, download or extract the following files into the same software directory:

- `PRSice.R`: R wrapper supplied by the PRSice repository
- `PRSice_win64.exe`: Windows PRSice executable

Linux and macOS users should replace the executable path below with the corresponding PRSice binary for their operating system.

## Target genotype data

Provide a PLINK binary dataset with a shared prefix:

```text
path/to/target_genotypes.bed
path/to/target_genotypes.bim
path/to/target_genotypes.fam
```

## GWAS summary statistics

Provide a tab-separated GWAS summary-statistics file. The example column mapping below expects the following source columns:

- `CHR`: chromosome
- `BP`: base-pair position
- `A1`: effect allele
- `A0`: non-effect allele
- `BETA`: effect-size estimate
- `P`: association p-value


## Phenotype and covariates

Provide a tab-separated file containing at least:

- `IID`: participant identifier matching the target genotype data
- `Phenotype`: phenotype used for PRS evaluation
- `Sex`: either `Male`/`Female` or numeric coding
- `Age`
- `PC1` to `PC10`: genetic principal components

The LIFE-Adult genotype and phenotype data used in the original analysis are access-controlled and cannot be downloaded from this repository.

# Analysis settings

Edit only the paths and study-specific settings in this section before running the notebook.

```{r settings}
trait <- "Diabetes"
study <- "Mahajan.NatGenet2018b.T2D.European"

# Set TRUE for a case-control phenotype and FALSE for a continuous phenotype.
is_binary_trait <- TRUE

# Number of threads used by PRSice.
n_threads <- 4

# PRSice-2 software files.
prsice_dir <- "path/to/PRSice"
prsice_r <- file.path(prsice_dir, "PRSice.R")
prsice_bin <- file.path(prsice_dir, "PRSice_win64.exe")

# Input files.
gwas_file <- "path/to/gwas_summary_statistics.tsv"
pheno_covars_file <- "path/to/phenotype_and_covariates.tsv"
target_prefix <- "path/to/target_genotypes"

# Output directory.
output_dir <- file.path("path/to/output", trait)
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

prepared_gwas_file <- file.path(
  output_dir,
  paste0(study, "_prsice.tsv")
)
pheno_file <- file.path(output_dir, "phenodata.tsv")
cov_file <- file.path(output_dir, "covardata.tsv")
out_prefix <- file.path(output_dir, "prsice")
```

# Prepare GWAS summary statistics

## Remove incomplete and duplicated variants

```{r prepare-sumstats}
sumstats <- fread(gwas_file)

sumstats <- sumstats[
  !is.na(CHR) &
  !is.na(BP) &
  !is.na(A1) &
  !is.na(A2) &
  !is.na(BETA) &
  !is.na(P)
]

sumstats[, duplicate_key := paste(
  CHR,
  BP,
  pmin(A1, A2),
  pmax(A1, A2),
  sep = ":"
)]

duplicated_variants <- sumstats[
  duplicated(duplicate_key) |
  duplicated(duplicate_key, fromLast = TRUE)
]

cat("Duplicated rows identified:", nrow(duplicated_variants), "\n")

sumstats <- sumstats[!duplicated(duplicate_key)]
sumstats[, duplicate_key := NULL]
```

## Match variant identifiers to the target BIM file

The target dataset used in the original analysis stored SNP identifiers as `CHR:BP:ALLELE1:ALLELE2`. Because allele order can differ between datasets, identifiers are constructed in both possible orders and matched against the target `.bim` file.

This step changes only the SNP identifier. It does not flip effect sizes or resolve strand-ambiguous variants. Allele harmonization and quality control should be completed before running this workflow.

```{r match-snp-ids}
bim <- fread(
  paste0(target_prefix, ".bim"),
  header = FALSE,
  col.names = c("CHR_BIM", "SNP_BIM", "CM", "BP_BIM", "BIM_A1", "BIM_A2")
)

sumstats[, SNP_original := paste(CHR, BP, A1, A2, sep = ":")]
sumstats[, SNP_reversed := paste(CHR, BP, A2, A1, sep = ":")]

sumstats[, matches_original := SNP_original %in% bim$SNP_BIM]
sumstats[, matches_reversed := SNP_reversed %in% bim$SNP_BIM]

sumstats[, SNP := fifelse(
  matches_original,
  SNP_original,
  fifelse(matches_reversed, SNP_reversed, SNP_original)
)]

n_overlap <- sum(sumstats$SNP %in% bim$SNP_BIM)

cat("GWAS variants after filtering:", nrow(sumstats), "\n")
cat("Variants in target BIM file:", nrow(bim), "\n")
cat("Variants with matching SNP IDs:", n_overlap, "\n")
cat("Overlap percentage:", round(100 * n_overlap / nrow(sumstats), 2), "%\n")
```

## Save the PRSice base file

Only the columns required for the PRSice analysis are written to disk.

```{r write-sumstats}
prsice_sumstats <- sumstats[, .(SNP, CHR, BP, A1, A2, BETA, P)]

fwrite(
  prsice_sumstats,
  prepared_gwas_file,
  sep = "\t",
  quote = FALSE,
  na = "NA"
)
```

# Prepare phenotype and covariate files

PRSice expects both phenotype and covariate files to contain `FID` and `IID`. This workflow uses `IID` for both identifiers.

For binary phenotypes, ensure that the phenotype coding is accepted by PRSice and by the downstream logistic-regression model. In this notebook, the downstream model assumes control/case coding as `0/1` or a compatible two-level outcome.

```{r prepare-pheno-covariates}
pheno_covars <- fread(pheno_covars_file)

required_pheno_columns <- c(
  "IID", "Phenotype", "Sex", "Age",
  paste0("PC", 1:10)
)
missing_pheno_columns <- setdiff(required_pheno_columns, names(pheno_covars))

if (length(missing_pheno_columns) > 0) {
  stop(
    "The phenotype/covariate file is missing: ",
    paste(missing_pheno_columns, collapse = ", ")
  )
}

# Convert character sex labels to PLINK-style numeric coding when necessary.
if (is.character(pheno_covars$Sex)) {
  pheno_covars[, Sex := fifelse(
    tolower(Sex) == "male", 1,
    fifelse(tolower(Sex) == "female", 2, NA_real_)
  )]
}

pheno <- pheno_covars[, .(
  FID = IID,
  IID = IID,
  Phenotype = Phenotype
)]

covariates <- pheno_covars[, c(
  list(FID = IID, IID = IID, Sex = Sex, Age = Age),
  mget(paste0("PC", 1:10))
)]

fwrite(pheno, pheno_file, sep = "\t", quote = FALSE, na = "NA")
fwrite(covariates, cov_file, sep = "\t", quote = FALSE, na = "NA")
```

# Run PRSice-2

The `--beta` option indicates that the `BETA` column contains beta coefficients rather than odds ratios. For an odds-ratio input column, use `--stat OR` and remove `--beta`.

```{r run-prsice}
prsice_args <- c(
  prsice_r,
  "--dir", prsice_dir,
  "--prsice", prsice_bin,
  "--base", prepared_gwas_file,
  "--target", target_prefix,
  "--pheno", pheno_file,
  "--pheno-col", "Phenotype",
  "--cov", cov_file,
  "--cov-col", paste(c("Age", "Sex", paste0("PC", 1:10)), collapse = ","),
  "--snp", "SNP",
  "--chr", "CHR",
  "--bp", "BP",
  "--a1", "A1",
  "--a2", "A2",
  "--stat", "BETA",
  "--pvalue", "P",
  "--beta",
  "--binary-target", ifelse(is_binary_trait, "T", "F"),
  "--thread", as.character(n_threads),
  "--out", out_prefix
)

rscript <- file.path(R.home("bin"), "Rscript")

prsice_output <- system2(
  command = rscript,
  args = prsice_args,
  stdout = TRUE,
  stderr = TRUE
)

cat(paste(prsice_output, collapse = "\n"))
```

# Inspect PRSice results

```{r read-results}
summary_file <- paste0(out_prefix, ".summary")

if (!file.exists(summary_file)) {
  stop("PRSice summary file was not created: ", summary_file)
}

prsice_summary <- fread(summary_file)
head(prsice_summary)
```

