---
title: "PRSice workflow"
output: html_document
---

# load libraries
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(glue)
library(data.table)
```

# define file paths
# sumstats need to have the columns SNP, CHR, BP, A1, A2, OR or BETA, P 
```{r}
trait <- "Diabetes"
study <- "Mahajan.NatGenet2018b.T2D.European"

#base_path   <- "R:/Doktoranden/MarieFreudenberg/analysen/PRSice/"
base_path   <- "H:/PRSice/"
prsice_dir  <- file.path(base_path, "PRSice_win64")
prsice_r    <- file.path(prsice_dir, "PRSice.R")
prsice_bin  <- file.path(prsice_dir, "PRSice_win64.exe")

# input files
target_prefix <- file.path(base_path, "LIFE_bed")
base_gwas     <- fread(file.path(glue("R:/Doktoranden/MarieFreudenberg/analysen/sumstats/{study}.tsv")))
pheno_covars_file <- fread(file.path(glue("R:/Doktoranden/MarieFreudenberg/analysen/Phenofiles/{trait}_covars.tsv")))

# output
out_prefix <- file.path(base_path, trait, "results", "prsice")
dir.create(dirname(out_prefix), showWarnings = FALSE, recursive = TRUE)
```

```{r}
sumstats <- (base_gwas)
head(sumstats)
```
# rename columns 
```{r}
sumstats <- sumstats %>%
  transmute(
   # SNP  = SNP,
    CHR  = chr,
    BP   = pos,
    A1   = a1,
    A2   = a0,
    BETA = beta,
    P    = p
  )
```

# remove doublicates from sumstats
```{r}
# allele-order-independent key
sumstats[, dup_key := paste(CHR, BP, pmin(A1, A2), pmax(A1, A2), sep=":")]

# inspect duplicated variants
dup_rows <- sumstats[duplicated(dup_key) | duplicated(dup_key, fromLast = TRUE)]
dup_rows[order(dup_key)]

# keep only one row per duplicated variant
sumstats <- sumstats[!duplicated(dup_key)]

# clean up
sumstats[, dup_key := NULL]
```

# naming in LIFE bim is SNP ID = CHR : BP : A2 : A1
# adjust sumstats accordingly
```{r}
# read bim file
bim <- fread(
  paste0(target_prefix, ".bim"),
  header = FALSE,
  col.names = c("CHR_bim", "SNP_bim", "CM", "BP_bim", "BIM_A1", "BIM_A2")
)

# create SNP IDs in both allele orders
sumstats[, SNP := paste(CHR, BP, A1, A2, sep=":")]
sumstats[, SNP_flipped := paste(CHR, BP, A2, A1, sep=":")]

# check which version matches the bim SNP IDs
sumstats[, match_original := SNP %in% bim$SNP_bim]
sumstats[, match_flipped  := SNP_flipped %in% bim$SNP_bim]

# prefer flipped ID if it matches, otherwise keep original
sumstats[, SNP := ifelse(match_flipped, SNP_flipped, SNP)]

# position needs to be an integer (no scientific writing)
sumstats[, BP := as.integer(BP)]

# save as tsv
fwrite(sumstats, file.path(glue(base_path, trait, "/{study}_prsice.tsv")), sep = "\t")
```
# check overlap
```{r}
# number of matches
n_overlap   <- sum((sumstats$SNP %in% bim$SNP_bim) | (sumstats$SNP_flipped %in% bim$SNP_bim))

cat("Total sumstats:   ", nrow(sumstats), "\n")
cat("Total bim SNPs:   ", nrow(bim), "\n")
cat("Overlap:          ", n_overlap, "\n")
```

```{r}
base_gwas     <- file.path(glue(base_path, trait, "/{study}_prsice.tsv"))
```

# Read in pheno and covar files, if they don' exist yet they need to be created using the next code chunk
```{r}
# define file paths
pheno_file <- file.path(base_path, trait, "phenodata.tsv")
cov_file <- file.path(base_path, trait, "covardata.tsv")
```

```{r}
# recode sex to 1 and 2 
pheno_covars_file[, Sex := fifelse(Sex == "Male", 1,
                            fifelse(Sex == "Female", 2, NA_real_))]

# create pheno file
pheno <- pheno_covars_file[, .(
  FID = IID,
  IID = IID,
  Phenotype = Phenotype
)]

# create covariate file
cov <- pheno_covars_file[, .(
  FID = IID,
  IID = IID,
  Sex = Sex,
  Age = Age,
  PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10
)]

# save as tab-separated files
fwrite(pheno, file.path(base_path, trait, "phenodata.tsv"), sep = "\t")
fwrite(cov, file.path(base_path, trait, "covardata.tsv"), sep = "\t")

pheno_file <- file.path(base_path, trait, "phenodata.tsv")
cov_file <- file.path(base_path, trait, "covardata.tsv")
```
# run PRSice
```{r}
args <- c(
  prsice_r,
  "--dir", prsice_dir,
  "--prsice", prsice_bin,
  "--base", base_gwas,
  "--target", target_prefix,
  "--pheno", pheno_file,
  "--pheno-col", "Phenotype",
  "--cov", cov_file,
  "--cov-col", "Age,Sex,PC1,PC2,PC3,PC4,PC5,PC6,PC7,PC8,PC9,PC10",
  "--stat", "BETA",
  "--beta",
  "--binary-target", "T", # for binary trait use T and for continuous trait use F
  "--thread", "4",
  "--out", out_prefix
)

cmd <- file.path(R.home("bin"), "Rscript")
res <- system2(cmd, args = args, stdout = TRUE, stderr = TRUE)
cat(paste(res, collapse = "\n"))
```


```{r read-results}
prsice_res <- read.table(
  paste0(out_prefix, ".summary"),
  header = TRUE,
  stringsAsFactors = FALSE
)

head(prsice_res)
```

# for binary traits AUC needs to be computed 
```{r}
library(pROC)

prs <- fread("H:/PRSice/Diabetes/results/prsice.best")
pheno <- fread("H:/PRSice/Diabetes/phenodata.tsv")
covar <- fread("H:/PRSice/Diabetes/covardata.tsv")

data <- merge(prs, pheno, by=c("FID","IID"))
data <- merge(data, covar, by=c("FID","IID"))

# PRS + covariates (full model)
model_full <- glm(
  Phenotype ~ PRS + Age + Sex + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10,
  data = data,
  family = binomial
)

# covariates only (base model)
model_null <- glm(
  Phenotype ~ Age + Sex + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10,
  data = data,
  family = binomial
)

auc_full <- roc(data$Phenotype, predict(model_full, type="response"))$auc
auc_null <- roc(data$Phenotype, predict(model_null, type="response"))$auc

auc_full
auc_null
```
# Odds ratio per SD increase in PRS
```{r}
# standardize PRS
data[, PRS_SD := as.numeric(scale(PRS))]

# logistic regression with standardized PRS
model_or <- glm(
  Phenotype ~ PRS_SD + Age + Sex +
    PC1 + PC2 + PC3 + PC4 + PC5 +
    PC6 + PC7 + PC8 + PC9 + PC10,
  data = data,
  family = binomial()
)

# extract coefficient and standard error
coef_prs <- coef(summary(model_or))["PRS_SD", "Estimate"]
se_prs   <- coef(summary(model_or))["PRS_SD", "Std. Error"]

# OR and 95% CI
OR <- exp(coef_prs)
CI_lower <- exp(coef_prs - 1.96 * se_prs)
CI_upper <- exp(coef_prs + 1.96 * se_prs)

cat("Odds Ratio per SD increase in PRS:", round(OR, 3), "\n")
cat("95% CI:", round(CI_lower, 3), "-", round(CI_upper, 3), "\n")
```

