---
title: "lassosum"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook constructs a polygenic score with `lassosum`, performs
split-validation to select the tuning parameters, and evaluates the resulting
score using linear or logistic regression.

The workflow follows the official lassosum repository:

<https://github.com/tshmak/lassosum>

# Required input

## GWAS summary statistics

A tab-separated file containing:

- `chr`: chromosome
- `pos`: base-pair position in hg38
- `a1`: effect allele
- `a0`: other allele
- `beta`: effect estimate
- `p`: association p-value
- `N`: GWAS sample size

## Genotype data

A PLINK BED dataset with one shared prefix:

```text
genotype_prefix.bed
genotype_prefix.bim
genotype_prefix.fam
```

The same dataset is used here as both the LD reference and test dataset, as in
the original workflow.

## Phenotype and covariate file

A tab-separated file containing:

```text
IID
Phenotype
Age
Sex
PC1-10
```

## Software

The `lassosum` R package and its dependencies are required.

# Load libraries

```{r}
library(RcppArmadillo)
library(Matrix)
library(data.table)
library(lassosum)
library(readr)
library(glue)
library(dplyr)
library(pROC)
library(DescTools)
```

# Set paths and analysis options

Change only the settings in this section.

```{r}
# Main project directory
path_prefix <- "path/to/project"

trait <- "Diabetes"
trait_type <- "binary"  # "continuous" or "binary"
study <- "Mahajan.NatGenet2018b.T2D.European"

# GWAS summary statistics
sumstats_file <- file.path(
  path_prefix,
  "sumstats",
  paste0(study, ".tsv")
)

# PLINK BED prefix without .bed, .bim, or .fam
genotype_prefix <- file.path(
  path_prefix,
  "path",
  "to",
  "genotype_data"
)

# Phenotype and covariate file
phenotype_file <- file.path(
  path_prefix,
  "Phenofiles",
  paste0(trait, "_covars.tsv")
)

# Output directory
output_dir <- file.path(
  path_prefix,
  "Lassosum",
  trait
)

dir.create(
  output_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

# LD block definition supplied by lassosum
LDblocks <- "EUR.hg38"
```

# Read and prepare summary statistics

```{r}
ss <- fread(
  sumstats_file
) %>%
  transmute(
    chr = chr,
    pos = pos,
    a1 = a1,
    a0 = a0,
    beta = beta,
    p = p,
    N = N
  )

ss$p <- as.numeric(ss$p)
ss$N <- as.numeric(ss$N)
ss$beta <- as.numeric(ss$beta)

# Replace exact p-values of 0 or 1
eps <- 1e-300

ss$p[ss$p <= 0] <- eps
ss$p[ss$p >= 1] <- 1 - eps
```

# Convert p-values to SNP–phenotype correlations

```{r}
cor <- p2cor(
  p = ss$p,
  n = ss$N,
  sign = sign(ss$beta)
)

# Remove variants for which the correlation could not be calculated
keep <- !is.na(cor)

cor <- cor[keep]
ss <- ss[keep, ]
```

# Run lassosum

```{r}
out <- lassosum.pipeline(
  cor = cor,
  chr = ss$chr,
  pos = ss$pos,
  A1 = ss$a1,
  A2 = ss$a0,
  ref.bfile = genotype_prefix,
  test.bfile = genotype_prefix,
  LDblocks = LDblocks
)
```

```{r}
lassosum_output_file <- file.path(
  output_dir,
  paste0(study, "_lassosum_out.rds")
)

saveRDS(
  out,
  lassosum_output_file
)
```

# Prepare phenotype and covariate input

`splitvalidate()` expects separate phenotype and covariate tables containing
both `FID` and `IID`.

```{r}
pheno_covars <- fread(
  phenotype_file
)

pheno <- pheno_covars[, .(
  FID = IID,
  IID = IID,
  Phenotype = Phenotype
)]

covar <- pheno_covars[, .(
  FID = IID,
  IID = IID,
  Sex = Sex,
  Age = Age,
  PC1,
  PC2,
  PC3,
  PC4,
  PC5,
  PC6,
  PC7,
  PC8,
  PC9,
  PC10
)]
```

# Split-validation

Split-validation selects the lassosum tuning parameters while reducing
overfitting when the same target dataset is used for model validation and score
evaluation.

```{r}
sv <- splitvalidate(
  out,
  pheno = pheno,
  covar = covar
)
```

# Create the PRS table

```{r}
pgs_vec <- sv$best.pgs

id_mat <- do.call(
  rbind,
  strsplit(
    names(pgs_vec),
    "_"
  )
)

colnames(id_mat) <- c(
  "FID",
  "IID"
)

prs_df <- data.frame(
  FID = id_mat[, "FID"],
  IID = id_mat[, "IID"],
  PRS = as.numeric(pgs_vec),
  row.names = NULL
)
```

# Merge PRS, phenotype, and covariates

```{r}
covariates <- c(
  "Age",
  "Sex",
  paste0("PC", 1:10)
)

analysis_data <- prs_df %>%
  left_join(
    pheno,
    by = c(
      "FID",
      "IID"
    )
  ) %>%
  left_join(
    covar,
    by = c(
      "FID",
      "IID"
    )
  ) %>%
  select(
    FID,
    IID,
    Phenotype,
    PRS,
    all_of(covariates)
  ) %>%
  filter(
    complete.cases(.)
  ) %>%
  mutate(
    Age = as.numeric(Age),
    Sex = as.factor(Sex),
    PRS_SD = as.numeric(
      scale(PRS)
    )
  )
```

# Define regression models

```{r}
base_formula <- as.formula(
  paste(
    "Phenotype ~",
    paste(
      covariates,
      collapse = " + "
    )
  )
)

full_formula <- as.formula(
  paste(
    "Phenotype ~",
    paste(
      c(
        "PRS_SD",
        covariates
      ),
      collapse = " + "
    )
  )
)
```

# Evaluate continuous traits

```{r}
if (trait_type == "continuous") {

  model_reduced <- lm(
    base_formula,
    data = analysis_data
  )

  model_full <- lm(
    full_formula,
    data = analysis_data
  )

  r2_base <- summary(
    model_reduced
  )$r.squared

  r2_full <- summary(
    model_full
  )$r.squared

  r2_prs <- r2_full - r2_base

  prs_row <- coef(
    summary(model_full)
  )["PRS_SD", ]

  prs_ci <- confint(
    model_full,
    "PRS_SD"
  )

  results_text <- glue(
    "
Base model R²:                {round(r2_base, 4)}
Full model R²:                {round(r2_full, 4)}
Incremental R² (PRS):         {round(r2_prs, 4)}

PRS effect per SD:
  Beta:                       {round(unname(prs_row['Estimate']), 4)}
  95% CI:                     [{round(prs_ci[1], 4)}, {round(prs_ci[2], 4)}]
  P-value:                    {signif(unname(prs_row['Pr(>|t|)']), 4)}
"
  )

  cat(
    results_text,
    "\n"
  )

  print(
    summary(model_full)
  )
}
```

# Evaluate binary traits

```{r}
if (trait_type == "binary") {

  y <- analysis_data$Phenotype

  if (
    is.factor(y) ||
      is.character(y)
  ) {

    analysis_data$Phenotype <-
      droplevels(
        as.factor(y)
      )

    if (
      nlevels(
        analysis_data$Phenotype
      ) != 2
    ) {
      stop(
        "For trait_type = 'binary', Phenotype must have exactly two levels."
      )
    }

  } else {

    unique_y <- sort(
      unique(y)
    )

    if (
      length(unique_y) != 2
    ) {
      stop(
        "For trait_type = 'binary', Phenotype must have exactly two values."
      )
    }

    analysis_data$Phenotype <-
      ifelse(
        y == unique_y[2],
        1,
        0
      )
  }

  model_reduced <- glm(
    base_formula,
    data = analysis_data,
    family = binomial()
  )

  model_full <- glm(
    full_formula,
    data = analysis_data,
    family = binomial()
  )

  probability_base <- predict(
    model_reduced,
    type = "response"
  )

  probability_full <- predict(
    model_full,
    type = "response"
  )

  roc_base <- pROC::roc(
    response = analysis_data$Phenotype,
    predictor = probability_base,
    quiet = TRUE
  )

  roc_full <- pROC::roc(
    response = analysis_data$Phenotype,
    predictor = probability_full,
    quiet = TRUE
  )

  auc_base <- as.numeric(
    pROC::auc(roc_base)
  )

  auc_full <- as.numeric(
    pROC::auc(roc_full)
  )

  auc_diff <- auc_full - auc_base

  nagelkerke_base <- unname(
    DescTools::PseudoR2(
      model_reduced,
      which = "Nagelkerke"
    )
  )

  nagelkerke_full <- unname(
    DescTools::PseudoR2(
      model_full,
      which = "Nagelkerke"
    )
  )

  nagelkerke_diff <-
    nagelkerke_full -
    nagelkerke_base

  prs_row <- coef(
    summary(model_full)
  )["PRS_SD", ]

  log_or <- unname(
    prs_row["Estimate"]
  )

  se_log_or <- unname(
    prs_row["Std. Error"]
  )

  p_value <- unname(
    prs_row["Pr(>|z|)"]
  )

  or_per_sd <- exp(log_or)

  or_ci <- exp(
    c(
      log_or - 1.96 * se_log_or,
      log_or + 1.96 * se_log_or
    )
  )

  results_text <- glue(
    "
Base model AUC:               {round(auc_base, 4)}
Full model AUC:               {round(auc_full, 4)}
Delta AUC (PRS):              {round(auc_diff, 4)}

Base model Nagelkerke R²:     {round(nagelkerke_base, 4)}
Full model Nagelkerke R²:     {round(nagelkerke_full, 4)}
Delta Nagelkerke R² (PRS):    {round(nagelkerke_diff, 4)}

PRS effect per SD:
  Log-odds beta:              {round(log_or, 4)}
  OR per SD:                  {round(or_per_sd, 4)}
  95% CI for OR:              [{round(or_ci[1], 4)}, {round(or_ci[2], 4)}]
  P-value:                    {signif(p_value, 4)}
"
  )

  cat(
    results_text,
    "\n"
  )

  print(
    summary(model_full)
  )
}
```

# Save scores and results

```{r}
write_tsv(
  analysis_data,
  file.path(
    output_dir,
    paste0(
      study,
      "_lassosum_scores.tsv"
    )
  )
)

saveRDS(
  sv,
  file.path(
    output_dir,
    paste0(
      study,
      "_splitvalidate.rds"
    )
  )
)
```
