---
title: "Untitled"
output: html_document
date: "2025-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# following https://github.com/tshmak/lassosum?tab=readme-ov-file

# install.packages("/path/to/downloaded_binary_file", repos=NULL)
# install.packages("R:/Doktoranden/MarieFreudenberg/analysen/Lassosum/lassosum_0.4.5.zip", repos=NULL)

# load libraries and define base path 
```{r}
.libPaths("H:/Rlibs")
library(RcppArmadillo)
library(Matrix)
library(data.table)
library(lassosum)
library(readr)
library(glue)
library(readxl)
library(dplyr)
library(pROC)
library(DescTools)

base_path <- "R:/Doktoranden/MarieFreudenberg/analysen/Lassosum/"
```

# Read summary statistics file (need positions on hg38)
```{r}
trait <- "Diabetes"
trait_type  <- "binary"   # "continuous" or "binary"
study <- "Mahajan.NatGenet2018b.T2D.European"
ss <- fread(file.path(glue("R:/Doktoranden/MarieFreudenberg/analysen/sumstats/{study}.tsv")))
head(ss)
nrow(ss)
```

# save sumstats as .tsv with correct column names needed for lassosum 
```{r}
ss <- ss %>%
  transmute(
    chr = chr,
    pos = pos,
    a1  = a1,
    a0  = a0,
    beta = beta,
    p    = p,
    N    = N
  )
fwrite(ss, file = file.path(glue(base_path, "{trait}/{study}_lassosum.tsv")))
```
# Specify the PLINK file stub of the reference panel
```{r}
ref.bfile <- file.path(base_path, "LIFE_bed")
```
# Specify the PLINK file stub of the test data (use same entire LIFE data for both LD reference and testing)
```{r}
test.bfile <- file.path(base_path, "LIFE_bed")
```
# Read LD region file
```{r}
LDblocks <- "EUR.hg38" 
```
# some p values are 0 or 1 which will be a problem later, increase/decrease the values a tiiny bit
```{r}
ss <- fread(file.path(glue(base_path, "{trait}/{study}_lassosum.tsv")))
ss$p    <- as.numeric(ss$p)
ss$N    <- as.numeric(ss$N)
ss$beta <- as.numeric(ss$beta)

sum(is.na(ss$p))
sum(is.na(ss$N))
sum(is.na(ss$beta))

table(ss$p == 0, useNA = "ifany")
table(ss$p == 1, useNA = "ifany")

eps <- 1e-300  # some tiny number

ss$p[ss$p <= 0] <- eps
ss$p[ss$p >= 1] <- 1 - eps
```

# To run lassosum, we need to input SNP-wise correlations with the phenotype. This can be converted from p-values via the p2cor function.
```{r}
cor <- p2cor(p = ss$p, n = ss$N, sign = sign(ss$beta))
```
# if there is a warning that some n are too small, drop those SNP
```{r}
keep <- !is.na(cor)
cor2 <- cor[keep]
ss  <- ss[keep, ]
```

# Running lassosum using standard pipeline (change to cor2 if cor function had warning)
```{r}
out <- lassosum.pipeline(cor=cor2, chr=ss$chr, pos=ss$pos, 
                         A1=ss$a1, A2=ss$a0, # A2 is not required but advised
                         ref.bfile=ref.bfile, test.bfile=test.bfile, 
                         LDblocks = LDblocks)
```
```{r}
saveRDS(out, file = file.path(glue(base_path, "{trait}/{study}_out.rds")))
```
```{r}
out <- readRDS(file.path(glue(base_path, "{trait}/{study}_out.rds")))
```

### Validation with phenotype ### 
# Read in pheno and covar files, if they don' exist yet they need to be created using the next code chunk
```{r}
pheno <- fread(file = file.path(glue(base_path, "{trait}/phenodata.tsv")))
covar <- fread(file = file.path(glue(base_path, "{trait}/covardata.tsv")))
```
```{r}
pheno_covars_file <- fread(file.path(glue("R:/Doktoranden/MarieFreudenberg/analysen/Phenofiles/{trait}_covars.tsv")))

# create pheno file
pheno_file <- pheno_covars_file[, .(
  FID = IID,
  IID = IID,
  Phenotype = Phenotype
)]

# create covariate file
cov_file <- pheno_covars_file[, .(
  FID = IID,
  IID = IID,
  Sex = Sex,
  Age = Age,
  PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10
)]

# save as tab-separated files
fwrite(pheno_file, file = file.path(base_path, trait, "phenodata.tsv"), sep = "\t")
fwrite(cov_file, file = file.path(base_path, trait, "covardata.tsv"), sep = "\t")

pheno <- fread(file = file.path(glue(base_path, "{trait}/phenodata.tsv")))
covar <- fread(file = file.path(glue(base_path, "{trait}/covardata.tsv")))
```

# Split-validation to avoid overfitting (one half for validation, other half for PGS calculation (PGS in the two halves are then standardised and stacked back together. This avoids overfitting due to the overlapping of the target and the validation dataset.))

```{r}
sv <- splitvalidate(out, pheno=pheno, covar=covar)
```
# run linear or logistic regression analysis
```{r}
outcome_var <- "Phenotype"
prs_var     <- "PRS"   # we will create this column from sv$best.pgs
covars      <- c("Age", "Sex", paste0("PC", 1:10))


# 1) PRS table from sv$best.pgs (names are "FID_IID")
pgs_vec <- sv$best.pgs
id_mat  <- do.call(rbind, strsplit(names(pgs_vec), "_"))
colnames(id_mat) <- c("FID", "IID")

prs_df <- data.frame(
  FID = id_mat[, "FID"],
  IID = id_mat[, "IID"],
  PRS = as.numeric(pgs_vec),
  row.names = NULL
)

# 2) Merge PRS with phenotype + covariates (assumes your pheno/covar have IID; add FID join if you have it)
analysis_data <- prs_df %>%
  left_join(pheno, by = "IID") %>%     # expects pheno has IID and Phenotype
  left_join(covar, by = "IID") %>%     # expects covar has IID and Age/Sex/PCs
  select(all_of(c("FID", "IID", outcome_var, prs_var, covars))) %>%
  as.data.frame()

# 3) Keep complete cases for all model variables
analysis_data <- analysis_data[complete.cases(analysis_data), ]

# 4) Ensure types
analysis_data$Age <- as.numeric(analysis_data$Age)
analysis_data$Sex <- as.factor(analysis_data$Sex)

# 5) Standardize PRS to PRS_SD (per SD effect)
analysis_data$PRS_SD <- as.numeric(scale(analysis_data[[prs_var]]))

# =========================
# Formulas (same as CS script)
# =========================
base_formula <- as.formula(
  paste(outcome_var, "~", paste(covars, collapse = " + "))
)
full_formula <- as.formula(
  paste(outcome_var, "~", paste(c("PRS_SD", covars), collapse = " + "))
)

# =========================
# Run models depending on trait type
# =========================
if (trait_type == "continuous") {

  # Linear regression
  model_reduced <- lm(base_formula, data = analysis_data)
  model_full    <- lm(full_formula, data = analysis_data)

  r2_base <- summary(model_reduced)$r.squared
  r2_full <- summary(model_full)$r.squared
  r2_prs  <- r2_full - r2_base

  prs_row <- coef(summary(model_full))["PRS_SD", ]
  prs_ci  <- confint(model_full, "PRS_SD")

  var_explained_txt <- glue("
Base model R²:                {round(r2_base, 4)}
Full model R²:                {round(r2_full, 4)}
Incremental R² (PRS):         {round(r2_prs, 4)}

PRS effect per SD:
  Beta:                       {round(unname(prs_row['Estimate']), 4)}
  95% CI:                     [{round(prs_ci[1], 4)}, {round(prs_ci[2], 4)}]
  P-value:                    {signif(unname(prs_row['Pr(>|t|)']), 4)}
")

  cat(var_explained_txt, "\n")
  print(summary(model_full))

  results <- list(
    trait_type   = trait_type,
    n            = nrow(analysis_data),
    base_model   = model_reduced,
    full_model   = model_full,
    r2_base      = r2_base,
    r2_full      = r2_full,
    r2_prs       = r2_prs,
    beta_per_sd  = unname(prs_row["Estimate"]),
    beta_ci_low  = prs_ci[1],
    beta_ci_high = prs_ci[2],
    p_value      = unname(prs_row["Pr(>|t|)"])
  )

} else if (trait_type == "binary") {

  # Make sure the outcome is binary
  y <- analysis_data[[outcome_var]]

  if (is.factor(y) || is.character(y)) {
    analysis_data[[outcome_var]] <- droplevels(as.factor(y))
    if (nlevels(analysis_data[[outcome_var]]) != 2) {
      stop("For trait_type = 'binary', the phenotype must have exactly 2 levels.")
    }
  } else {
    uniq_y <- sort(unique(y))
    if (length(uniq_y) != 2) {
      stop("For trait_type = 'binary', the phenotype must have exactly 2 values.")
    }
    # Recode numeric/logical outcomes to 0/1 using the higher value as the event
    analysis_data[[outcome_var]] <- ifelse(y == uniq_y[2], 1, 0)
  }

  # Logistic regression
  model_reduced <- glm(base_formula, data = analysis_data, family = binomial())
  model_full    <- glm(full_formula, data = analysis_data, family = binomial())

  # Predicted probabilities
  prob_base <- predict(model_reduced, type = "response")
  prob_full <- predict(model_full, type = "response")

  # AUC
  roc_base <- pROC::roc(response = analysis_data[[outcome_var]], predictor = prob_base, quiet = TRUE)
  roc_full <- pROC::roc(response = analysis_data[[outcome_var]], predictor = prob_full, quiet = TRUE)

  auc_base <- as.numeric(pROC::auc(roc_base))
  auc_full <- as.numeric(pROC::auc(roc_full))
  auc_diff <- auc_full - auc_base

  # Nagelkerke R²
  nag_base <- unname(DescTools::PseudoR2(model_reduced, which = "Nagelkerke"))
  nag_full <- unname(DescTools::PseudoR2(model_full, which = "Nagelkerke"))
  nag_diff <- nag_full - nag_base

  # PRS effect per SD
  prs_row   <- coef(summary(model_full))["PRS_SD", ]
  log_or    <- unname(prs_row["Estimate"])
  se_log_or <- unname(prs_row["Std. Error"])
  p_value   <- unname(prs_row["Pr(>|z|)"])

  or_per_sd <- exp(log_or)
  or_ci     <- exp(c(log_or - 1.96 * se_log_or, log_or + 1.96 * se_log_or))
  odds_increase_pct <- (or_per_sd - 1) * 100

  logistic_txt <- glue("
Base model AUC:               {round(auc_base, 4)}
Full model AUC:               {round(auc_full, 4)}
Delta AUC (PRS):              {round(auc_diff, 4)}

Base model Nagelkerke R²:     {round(nag_base, 4)}
Full model Nagelkerke R²:     {round(nag_full, 4)}
Delta Nagelkerke R² (PRS):    {round(nag_diff, 4)}

PRS effect per SD:
  Log-odds beta:              {round(log_or, 4)}
  OR per SD:                  {round(or_per_sd, 4)}
  95% CI for OR:              [{round(or_ci[1], 4)}, {round(or_ci[2], 4)}]
  % increase in odds / SD:    {round(odds_increase_pct, 2)}%
  P-value:                    {signif(p_value, 4)}
")

  cat(logistic_txt, "\n")
  print(summary(model_full))

  results <- list(
    trait_type         = trait_type,
    n                  = nrow(analysis_data),
    base_model         = model_reduced,
    full_model         = model_full,
    auc_base           = auc_base,
    auc_full           = auc_full,
    auc_diff           = auc_diff,
    nagelkerke_base    = nag_base,
    nagelkerke_full    = nag_full,
    nagelkerke_diff    = nag_diff,
    log_or_per_sd      = log_or,
    or_per_sd          = or_per_sd,
    or_ci_low          = or_ci[1],
    or_ci_high         = or_ci[2],
    odds_increase_pct  = odds_increase_pct,
    p_value            = p_value
  )

} else {
  stop("trait_type must be either 'continuous' or 'binary'.")
}






