---
title: "GWAS Manhattan Plot"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This notebook creates a Manhattan plot from GWAS results and optionally
highlights variants.

## Required input

The GWAS results file must contain:

```text
chr
pos
SNP
p
```

The optional SNP input file must contain one SNP identifier per line.

The SNP identifiers must match the format used in the GWAS `SNP` column, for
example:

```text
3:150211647
```

# Load libraries

```{r}
library(qqman)
library(readr)
library(dplyr)
library(stringr)
library(purrr)
```

# Set paths and analysis options

```{r}
# Main project directory
path_prefix <- "path/to/project"

trait <- "BMI"

# GWAS results
gwas_file <- file.path(
  path_prefix,
  "GWAS",
  paste0(
    "plinkresults.",
    trait,
    ".glm.linear_harmonized_minimal.txt.gz"
  )
)

# Set to TRUE to highlight SNPs from the input list
highlight_snps <- TRUE

# Text file containing one SNP identifier per line
snp_list_file <- file.path(
  path_prefix,
  "GWAS",
  "snps_of_interest.txt"
)
```

# Read and prepare the GWAS results

```{r}
df <- read_tsv(
  gwas_file,
  show_col_types = FALSE
)

df <- df %>%
  filter(
    !chr %in% c(
      "X",
      "Y",
      "MT"
    )
  ) %>%
  mutate(
    chr = as.numeric(chr),
    pos = as.numeric(pos),
    p = as.numeric(p)
  ) %>%
  filter(
    !is.na(chr),
    !is.na(pos),
    !is.na(p),
    p <= 1
  )
```

# Replace zero p-values for plotting

Zero p-values cannot be transformed with `-log10()`. They are replaced only in
the plotting column with a value slightly smaller than the smallest non-zero
p-value.

```{r}
min_nonzero <- min(
  df$p[df$p > 0],
  na.rm = TRUE
)

p_floor <- max(
  .Machine$double.xmin,
  min_nonzero / 10
)

df <- df %>%
  mutate(
    p_plot = if_else(
      p <= 0,
      p_floor,
      p
    )
  )
```

# Manhattan plot without highlighted SNPs

```{r}
manhattan(
  df,
  chr = "chr",
  bp = "pos",
  snp = "SNP",
  p = "p_plot"
)
```

# Read SNPs to highlight

This section runs only when `highlight_snps` is set to `TRUE`.

```{r}
if (highlight_snps) {

  snps_of_interest <- read_lines(
    snp_list_file
  ) %>%
    str_trim() %>%
    discard(~ .x == "") %>%
    unique()
}
```

# Manhattan plot with highlighted variants

```{r}
if (highlight_snps) {

  manhattan(
    df,
    chr = "chr",
    bp = "pos",
    snp = "SNP",
    p = "p_plot",
    highlight = snps_of_interest
  )
}
```

# Manhattan plot with annotated associations

```{r}
manhattan(
  df,
  chr = "chr",
  bp = "pos",
  snp = "SNP",
  p = "p_plot",
  annotatePval = 0.01
)
```