---
title: "Create hg38 Score Files from PGS Catalog Downloads"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This notebook creates a score file from a PGS Catalog download. If the score is
provided in hg19/GRCh37 coordinates, the variants are first converted to
hg38/GRCh38 coordinates using the UCSC liftOver website.

## Required files

- A scoring file downloaded from the PGS Catalog:
  <https://www.pgscatalog.org/>
- The `.pvar` file belonging to the target PLINK 2 genotype dataset.
- For hg19 score files, the UCSC liftOver website:
  <https://genome.ucsc.edu/cgi-bin/hgLiftOver>

# Load libraries

```{r}
library(readr)
library(dplyr)
library(glue)
library(stringr)
```

# Specify paths and score

Change only the path prefix and filenames in this section.

```{r}
# Main project path
path_prefix <- "path/to/project"

# PGS Catalog score identifier
file <- "PGS005204"

# Downloaded PGS Catalog scoring file
PGSfile <- file.path(path_prefix, "LIFE_PRS", paste0(file, ".txt"))

# Variant file for the target genotype dataset
pvar_file <- file.path(path_prefix, "path", "to", "genotype_data.pvar")

pvar <- read_tsv(
  pvar_file,
  comment = "##",
  col_names = TRUE
)
```

# Read the PGS file

PGS Catalog files contain metadata lines beginning with `#`. The function
automatically counts and skips these lines.

```{r}
read_prs_file <- function(PGSfile) {
  # Read the first 50 lines
  header_lines <- readLines(PGSfile, n = 50)

  # Count lines beginning with "#"
  skip_lines <- sum(grepl("^#", header_lines))

  # Read the scoring table
  read_delim(PGSfile, skip = skip_lines)
}

pgs <- read_prs_file(PGSfile)
```

# Convert hg19 coordinates to BED format

Run this section if the downloaded score file uses hg19/GRCh37 coordinates.

BED files use zero-based, half-open coordinates.

```{r}
bed <- pgs %>%
  mutate(
    chrom = paste0("chr", chr_name),
    start = chr_position - 1,
    end = chr_position,
    id = row_number()
  ) %>%
  select(chrom, start, end, id)

bed_path <- file.path(
  path_prefix,
  "LIFE_PRS",
  paste0(file, "_hg19.bed")
)

write_tsv(bed, bed_path, col_names = FALSE)
```

Upload the BED file to the UCSC liftOver website and convert it from hg19 to
hg38:

<https://genome.ucsc.edu/cgi-bin/hgLiftOver>

Save the converted file as:

```text
path/to/project/LIFE_PRS/<PGS_ID>_hg38.bed
```

# Convert the lifted BED file back to a scoring table

```{r}
bed_path <- file.path(
  path_prefix,
  "LIFE_PRS",
  paste0(file, "_hg38.bed")
)

lifted <- read_tsv(bed_path, col_names = FALSE) %>%
  select(1:4) %>%
  rename(
    chr_hg38 = X1,
    start_hg38 = X2,
    end_hg38 = X3,
    row_id = X4
  ) %>%
  mutate(
    pos_hg38 = end_hg38,
    row_id = as.character(row_id)
  )

pgs <- pgs %>%
  mutate(row_id = as.character(row_number()))

pgs_lifted <- inner_join(pgs, lifted, by = "row_id")
```

# Rename SNP IDs to match the target genotype data

The target variant IDs are assumed to use the following format:

```text
chr:position:allele1:allele2
```

```{r}
scorefile <- pgs_lifted %>%
  mutate(
    chr_name = gsub("^chr", "", chr_hg38),
    chr_name = paste0("chr", chr_name),
    chr_position = pos_hg38,
    SNP = paste0(
      chr_name,
      ":",
      chr_position,
      ":",
      effect_allele,
      ":",
      other_allele
    )
  ) %>%
  select(
    SNP,
    A1 = effect_allele,
    BETA = effect_weight
  )
```

# Create flipped SNP IDs

The order of the two alleles in some SNP IDs must be reversed to match the
variant IDs in the target `.pvar` file.

```{r}
scorefile <- scorefile %>%
  mutate(
    chr = str_extract(SNP, "^chr[0-9XY]+"),
    pos = str_extract(SNP, "(?<=:)[0-9]+"),
    ref = str_extract(SNP, ":[ACGT]+(?=:)") %>%
      str_remove("^:"),
    alt = str_extract(SNP, ":[ACGT]+$") %>%
      str_remove("^:"),
    SNP_flipped = paste0(
      chr,
      ":",
      pos,
      ":",
      alt,
      ":",
      ref
    )
  )
```

# Match SNP IDs to the target `.pvar` file

Check which of the two IDs occurs in the `.pvar` file. Use the flipped ID when
it matches; otherwise retain the original ID.

```{r}
scorefile <- scorefile %>%
  mutate(
    match_flipped = SNP_flipped %in% pvar$ID,
    match_original = SNP %in% pvar$ID,
    SNP = if_else(match_flipped, SNP_flipped, SNP)
  ) %>%
  select(
    -SNP_flipped,
    -match_flipped,
    -match_original
  )
```

# Save the final score file

```{r}
final_scorefile <- scorefile %>%
  select(SNP, A1, BETA) %>%
  distinct(SNP, .keep_all = TRUE)

output_file <- file.path(
  path_prefix,
  "LIFE_PRS",
  paste0(file, "_scorefile.txt")
)

write_tsv(final_scorefile, output_file)
```
