---
title: "Untitled"
output: html_document
date: "2025-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
--------------------------------------------------------------------------------------------------------
# create scorefiles from PGS catalog downloads (usually in hg19 so they need to be lifted to hg38 first)
--------------------------------------------------------------------------------------------------------

# load libraries 
```{r}
library(readr)
library(dplyr)
library(glue)
library(stringr)
```

# Specify trait and file
```{r}
file <- "PGS005204"
PGSfile <- glue("R:/Doktoranden/MarieFreudenberg/analysen/LIFE_PRS/{file}.txt")
pvar <- read_tsv("R:/Doktoranden/MarieFreudenberg/genetik_LIFE_Adult/LIFE-Adult_TOPMed_minimal_filtered_chr1to23.pvar",
                 comment = "##", col_names = TRUE)
```

# Read PGS file 
```{r}
read_prs_file <- function(PGSfile) {
  # Read first 50 lines
  header_lines <- readLines(PGSfile, n = 50)
  
  # Count how many start with "#"
  skip_lines <- sum(grepl("^#", header_lines))
  
  # Read file with auto-calculated skip
  read_delim(PGSfile, skip = skip_lines)
}
pgs <- read_prs_file(PGSfile)
```
# if the file is in hg19 it needs to be converted to 38
# change file to bed format (0-based, half-open)
```{r}
bed <- pgs %>%
  mutate(
    chrom = paste0("chr", chr_name),
    start = chr_position - 1,
    end = chr_position,
    id = row_number()
  ) %>%
  select(chrom, start, end, id)

bed_path <- glue("R:/Doktoranden/MarieFreudenberg/analysen/LIFE_PRS/{file}_hg19.bed")
write_tsv(bed, bed_path, col_names = FALSE)
```

# go to this website and convert bed file from hg19 to hg38 and manually move and rename the file to {file}_hg38.bed
# https://genome.ucsc.edu/cgi-bin/hgLiftOver

# bed file needs to be converted back to a scorefile
```{r}
bed_path <- glue("R:/Doktoranden/MarieFreudenberg/analysen/LIFE_PRS/{file}_hg38.bed")
lifted <- read_tsv(bed_path, col_names = FALSE) %>%
  select(1:4) %>%
  rename(chr_hg38 = X1, start_hg38 = X2, end_hg38 = X3, row_id = X4) %>%
  mutate(
    pos_hg38 = end_hg38,
    row_id = as.character(row_id)
  )

pgs <- pgs %>%
  mutate(row_id = as.character(row_number()))

pgs_lifted <- inner_join(pgs, lifted, by = "row_id")
```

# rename SNP ID to match LIFE-Adult genodata
```{r}
scorefile <- pgs_lifted %>%
  mutate(
    chr_name = gsub("^chr", "", chr_hg38),   # clean up chromosome label
    chr_name = paste0("chr", chr_name),      # ensure consistent format
    chr_position = pos_hg38,
    SNP = paste0(chr_name, ":", chr_position, ":", effect_allele, ":", other_allele)
  ) %>%
  select(SNP, A1 = effect_allele, BETA = effect_weight)
```

# the order of the two alleles in some SNP IDs need to be flipped 
# create a s second ID column for the flipped SNP IDs
```{r}
scorefile <- scorefile %>%
  mutate(
    chr   = str_extract(SNP, "^chr[0-9XY]+"),
    pos   = str_extract(SNP, "(?<=:)[0-9]+"),
    ref   = str_extract(SNP, ":[ACGT]+(?=:)") %>% str_remove("^:"),
    alt   = str_extract(SNP, ":[ACGT]+$") %>% str_remove("^:"),
    SNP_flipped = paste0(chr, ":", pos, ":", alt, ":", ref)
  )
```

# check which of these two IDs match with the pvar file and keep only the matching ID (if neither match keep original) 
```{r}
scorefile <- scorefile %>%
  mutate(
    match_flipped  = SNP_flipped %in% pvar$ID,
    match_original = SNP %in% pvar$ID,
    SNP = if_else(match_flipped, SNP_flipped, SNP)  # prefer flipped when it matches
  ) %>%
  select(-SNP_flipped, -match_flipped, -match_original) 
```

# create a final scorefile and save as tsv
```{r}
final_scorefile <- scorefile %>%
  select(SNP, A1, BETA) %>%
  distinct(SNP, .keep_all = TRUE)

write_tsv(final_scorefile, glue("R:/Doktoranden/MarieFreudenberg/analysen/LIFE_PRS/{file}_scorefile.txt"))
```
