---
title: "PRS Regression Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This notebook evaluates the association between a standardized polygenic risk
score and a phenotype.

Set `outcome_type` to:

- `"continuous"` to run the original linear-regression workflow;
- `"binary"` to run the original logistic-regression workflow.

## Required files

- A phenotype and covariate file containing:
  `IID`, `Phenotype`, `Age`, `Sex`, and `PC1` to `PC10`.
- A standardized PRS file containing:
  `IID` and `SCORE_standardized`.

# 1. Load packages

```{r packages}
library(readr)
library(dplyr)
library(glue)
library(ggplot2)
library(tibble)
library(knitr)
library(pROC)
```

# 2. Analysis settings

Change only the settings in this section.

```{r settings}
# Main project path
path_prefix <- "path/to/project"

# Trait and PGS identifier
trait <- "Height"
pgs_file <- "PGS002804"

# Type of phenotype:
# "continuous" for linear regression
# "binary" for logistic regression
outcome_type <- "continuous"

phenotype_file <- file.path(
  path_prefix,
  "Phenofiles",
  paste0(trait, "_covars.tsv")
)

prs_file <- file.path(
  path_prefix,
  "LIFE_PRS",
  paste0(pgs_file, "_output_standardized.tsv")
)

score_col <- "SCORE_standardized"

covariates <- c(
  "Age",
  "Sex",
  paste0("PC", 1:10)
)
```

# 3. Load and merge data

```{r load-data}
phenotype_data <- read_tsv(
  phenotype_file,
  show_col_types = FALSE
)

prs_data <- read_tsv(
  prs_file,
  show_col_types = FALSE
)

data <- phenotype_data %>%
  left_join(prs_data, by = "IID")
```

# 4. Prepare analysis data

```{r prepare-data}
model_variables <- c(
  "IID",
  "Phenotype",
  score_col,
  covariates
)

data_model <- data %>%
  select(all_of(model_variables)) %>%
  filter(complete.cases(.)) %>%
  mutate(Sex = as.factor(Sex))

if (outcome_type == "binary") {
  data_model <- data_model %>%
    filter(Phenotype %in% c(0, 1)) %>%
    mutate(Phenotype = as.integer(Phenotype))
}
```

# 5. Define model formulas

```{r formulas}
base_formula <- as.formula(
  paste(
    "Phenotype ~",
    paste(covariates, collapse = " + ")
  )
)

full_formula <- as.formula(
  paste(
    "Phenotype ~",
    paste(c(score_col, covariates), collapse = " + ")
  )
)
```

# 6. Linear regression

```{r linear-regression}
if (outcome_type == "continuous") {

  model_reduced <- lm(
    base_formula,
    data = data_model
  )

  model <- lm(
    full_formula,
    data = data_model
  )

  r2_full <- summary(model)$r.squared
  r2_covars <- summary(model_reduced)$r.squared
  r2_prs <- r2_full - r2_covars

  var_explained_txt <- glue(
    "
Base model R²:                {round(r2_covars, 4)}
Full model R²:                {round(r2_full, 4)}
Incremental R²:               {round(r2_prs, 4)}
"
  )

  cat(var_explained_txt, "\n")
}
```

## PRS effect per standard deviation

```{r linear-effect}
if (outcome_type == "continuous") {

  coef_table <- summary(model)$coefficients

  beta_per_sd <- coef_table[
    score_col,
    "Estimate"
  ]

  se_per_sd <- coef_table[
    score_col,
    "Std. Error"
  ]

  p_value <- coef_table[
    score_col,
    "Pr(>|t|)"
  ]

  ci_per_sd <- confint(
    model,
    score_col,
    level = 0.95
  )

  beta_txt <- glue(
    "
PRS effect per SD:
  Beta per SD:                 {round(beta_per_sd, 4)} {phenotype_unit}
  95% CI:                      [{round(ci_per_sd[1], 4)}, {round(ci_per_sd[2], 4)}] {phenotype_unit}
  Standard error:              {round(se_per_sd, 4)}
  P-value:                     {format.pval(p_value, digits = 4, eps = 1e-300)}
"
  )

  cat(beta_txt, "\n")
}
```

## Predicted phenotype

```{r linear-prediction}
if (outcome_type == "continuous") {

  phenotype_mean <- mean(
    data_model$Phenotype,
    na.rm = TRUE
  )

  effect_size <- coef(model)[score_col]

  data_model <- data_model %>%
    mutate(
      predicted = phenotype_mean +
        .data[[score_col]] * effect_size,
      predicted_full = predict(
        model,
        newdata = data_model
      )
    )
}
```

## Predicted versus measured phenotype

```{r linear-predicted-plot}
if (outcome_type == "continuous") {

  ggplot(
    data_model,
    aes(
      x = predicted_full,
      y = Phenotype
    )
  ) +
    geom_point(
      alpha = 0.4,
      color = "steelblue"
    ) +
    geom_smooth(
      method = "lm",
      color = "darkred"
    ) +
    labs(
      title = glue(
        "Predicted vs Measured {trait}"
      ),
      x = "Predicted Phenotype",
      y = "Measured Phenotype"
    ) +
    theme_minimal()
}
```

## PRS effect on phenotype

```{r linear-prs-plot}
if (outcome_type == "continuous") {

  ggplot(
    data_model,
    aes(
      x = .data[[score_col]],
      y = Phenotype
    )
  ) +
    geom_point(alpha = 0.3) +
    geom_smooth(
      method = "lm",
      formula = y ~ x,
      color = "blue"
    ) +
    labs(
      title = glue(
        "Effect of {trait} PRS on Phenotype"
      ),
      x = "Standardized PRS",
      y = "Phenotype"
    ) +
    theme_minimal()
}
```

# 6. Logistic regression

This section runs only when:

```r
outcome_type <- "binary"
```

## Fit logistic regression models

```{r logistic-models}
if (outcome_type == "binary") {

  m_null <- glm(
    Phenotype ~ 1,
    data = data_model,
    family = binomial()
  )

  m_base <- glm(
    base_formula,
    data = data_model,
    family = binomial()
  )

  m_full <- glm(
    full_formula,
    data = data_model,
    family = binomial()
  )
}
```

## PRS effect per standard deviation

Because the PRS is standardized, the odds ratio represents the increase in odds
per one-standard-deviation higher PRS.

```{r logistic-effect}
if (outcome_type == "binary") {

  prs_coef <- summary(m_full)$coefficients[
    score_col,
  ]

  prs_beta <- prs_coef["Estimate"]
  prs_se <- prs_coef["Std. Error"]
  prs_p <- prs_coef["Pr(>|z|)"]

  prs_or <- exp(prs_beta)

  prs_or_ci_low <- exp(
    prs_beta - 1.96 * prs_se
  )

  prs_or_ci_high <- exp(
    prs_beta + 1.96 * prs_se
  )

  prs_effect_table <- tibble(
    term = score_col,
    beta = prs_beta,
    SE = prs_se,
    OR_per_SD = prs_or,
    CI_low = prs_or_ci_low,
    CI_high = prs_or_ci_high,
    p_value = prs_p
  )

  kable(
    prs_effect_table,
    digits = 4
  )
}
```

## AUC comparison

```{r logistic-auc}
if (outcome_type == "binary") {

  p_base <- predict(
    m_base,
    type = "response"
  )

  p_full <- predict(
    m_full,
    type = "response"
  )

  roc_base <- roc(
    data_model$Phenotype,
    p_base,
    quiet = TRUE
  )

  roc_full <- roc(
    data_model$Phenotype,
    p_full,
    quiet = TRUE
  )

  auc_base <- as.numeric(
    auc(roc_base)
  )

  auc_full <- as.numeric(
    auc(roc_full)
  )

  delta_auc <- auc_full - auc_base

  ci_auc_base <- ci.auc(roc_base)
  ci_auc_full <- ci.auc(roc_full)

  auc_test <- roc.test(
    roc_base,
    roc_full,
    paired = TRUE,
    method = "delong"
  )

  auc_results <- tibble(
    model = c(
      "Base",
      "Full: Base + PRS",
      "Difference"
    ),
    AUC = c(
      auc_base,
      auc_full,
      delta_auc
    ),
    CI_low = c(
      ci_auc_base[1],
      ci_auc_full[1],
      NA
    ),
    CI_high = c(
      ci_auc_base[3],
      ci_auc_full[3],
      NA
    ),
    p_value_DeLong = c(
      NA,
      NA,
      auc_test$p.value
    )
  )

  kable(
    auc_results,
    digits = 4
  )
}
```

## ROC plot

```{r logistic-roc-plot}
if (outcome_type == "binary") {

  plot(
    roc_base,
    col = "black",
    lwd = 2,
    main = glue(
      "ROC Curve: {trait}"
    )
  )

  plot(
    roc_full,
    col = "blue",
    lwd = 2,
    add = TRUE
  )

  legend(
    "bottomright",
    legend = c(
      paste0(
        "Base model = ",
        round(auc_base, 3)
      ),
      paste0(
        "Full model = ",
        round(auc_full, 3)
      )
    ),
    col = c("black", "blue"),
    lwd = 2
  )
}
```

## PRS distribution by phenotype

```{r logistic-boxplot}
if (outcome_type == "binary") {

  ggplot(
    data_model,
    aes(
      x = factor(Phenotype),
      y = .data[[score_col]]
    )
  ) +
    geom_boxplot(outlier.alpha = 0.3) +
    geom_jitter(
      width = 0.15,
      alpha = 0.2
    ) +
    labs(
      title = glue(
        "{trait} PRS by phenotype status"
      ),
      x = "Phenotype",
      y = "Standardized PRS"
    ) +
    theme_minimal()
}
```

