---
title: "Untitled"
output: html_document
date: "2025-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(readr)
library(dplyr)
library(glue)
```

# Specify trait and file
```{r}
trait <- "Diabetes"
PGSfile <- "R:/Doktoranden/MarieFreudenberg/analysen/1kGp3_PRS/PGS000014.txt.gz"
```
```{r}
pvar <- read_tsv("R:/Doktoranden/MarieFreudenberg/daten/1kG_ph3/1kgALL_ph3_hg38_maf001_uniqueSNPIDs.pvar",
                 comment = "##", col_names = TRUE)
```

# Read PGS file 
```{r}
read_prs_file <- function(PGSfile) {
  # Read first 50 lines
  header_lines <- readLines(PGSfile, n = 50)
  
  # Count how many start with "#"
  skip_lines <- sum(grepl("^#", header_lines))
  
  # Read file with auto-calculated skip
  read_delim(PGSfile, skip = skip_lines)
}
pgs <- read_prs_file(PGSfile)
```

# change file to bed format (0-based, half-open)
```{r, eval = FALSE}
bed <- pgs %>%
  mutate(
    chrom = paste0("chr", chr_name),
    start = chr_position - 1,
    end = chr_position,
    id = row_number()
  ) %>%
  select(chrom, start, end, id)

bed_path <- glue("R:/Doktoranden/MarieFreudenberg/analysen/1kGp3_PRS/{trait}_pgs_hg19.bed")
write_tsv(bed, bed_path, col_names = FALSE)
```

# go to this website and convert bed file from hg19 to hg38 and manually move and rename the file to {trait}_pgs_hg38.bed
# https://genome.ucsc.edu/cgi-bin/hgLiftOver

# bed file needs to be converted back to a scorefile
```{r}
bed_path <- glue("R:/Doktoranden/MarieFreudenberg/analysen/1kGp3_PRS/{trait}_pgs_hg38.bed")
lifted <- read_tsv(bed_path, col_names = FALSE) %>%
  select(1:4) %>%
  rename(chr_hg38 = X1, start_hg38 = X2, end_hg38 = X3, row_id = X4) %>%
  mutate(
    pos_hg38 = end_hg38,
    row_id = as.character(row_id)
  )

pgs <- pgs %>%
  mutate(row_id = as.character(row_number()))

pgs_lifted <- inner_join(pgs, lifted, by = "row_id")
```

```{r}
scorefile <- pgs_lifted %>%
  mutate(
    chr_position = pos_hg38,
    SNP = paste0(chr_name, ":", chr_position, ":", effect_allele, ":", other_allele)
  ) %>%
  select(SNP, A1 = effect_allele, BETA = effect_weight)
```

# OPTIONAL: Check how many original SNP ID match, how many flipped IDs would match and how many positions overlap?
```{r, eval = FALSE}
scorefile_info <- scorefile %>%
  mutate(
    chr = str_extract(SNP, "^[0-9XY]+"),
    pos = str_extract(SNP, "(?<=:)[0-9]+") %>% as.integer(),
    ref = str_extract(SNP, ":[ACGT]+(?=:)") %>% str_remove("^:"),
    alt = str_extract(SNP, ":[ACGT]+$") %>% str_remove("^:"),
    POS_key = paste0(chr, ":", pos),
    SNP_flipped = paste0(chr, ":", pos, ":", alt, ":", ref)
  )

pvar <- pvar %>%
  mutate(
    chr = as.character(`#CHROM`),
    POS_key = paste0(chr, ":", POS)
  )

n_total <- nrow(scorefile_info)
n_original_matched <- sum(scorefile_info$SNP %in% pvar$ID)
n_flipped_matched  <- sum(scorefile_info$SNP_flipped %in% pvar$ID)
n_total_matched    <- sum(scorefile_info$SNP %in% pvar$ID | scorefile_info$SNP_flipped %in% pvar$ID)
n_unmatched        <- n_total - n_total_matched

pos_overlap <- intersect(scorefile_info$POS_key, pvar$POS_key)
n_pos_overlap <- length(pos_overlap)

# Print summary report
cat(glue("
Total SNPs in scorefile:     {n_total}
Matched by original ID:      {n_original_matched}
Matched by flipped ID:       {n_flipped_matched}
Total SNPs matched:          {n_total_matched}
Unmatched SNPs:              {n_unmatched}
Positions matched (POS_key): {n_pos_overlap}
Position match rate:         {round(n_pos_overlap / n_total * 100, 2)}%
"))
```

# create a s second ID column for the flipped SNP IDs
```{r}
scorefile <- scorefile %>%
  mutate(
    chr   = str_extract(SNP, "^[0-9XY]+"),
    pos   = str_extract(SNP, "(?<=:)[0-9]+"),
    ref   = str_extract(SNP, ":[ACGT]+(?=:)") %>% str_remove("^:"),
    alt   = str_extract(SNP, ":[ACGT]+$") %>% str_remove("^:"),
    SNP_flipped = paste0(chr, ":", pos, ":", alt, ":", ref)
  )
```

# check which of these two IDs match with the pvar file and keep only the matching ID to create a new scorefile (if neither match keep original) 
```{r}
scorefile <- scorefile %>%
  mutate(
    in_pvar_original = SNP %in% pvar$ID,
    in_pvar_flipped  = SNP_flipped %in% pvar$ID,
    
    SNP_final = case_when(
      in_pvar_original ~ SNP,
      in_pvar_flipped  ~ SNP_flipped,
      TRUE ~ SNP  # fallback to original
    ),
    
    was_flipped = in_pvar_flipped & !in_pvar_original,
    was_matched = in_pvar_original | in_pvar_flipped
  )
```

# OPTIONAL: print how many of the original IDs were actually replaced by flipped ID
```{r}
total <- nrow(scorefile)
n_flipped <- sum(scorefile$was_flipped)
n_matched <- sum(scorefile$was_matched)
n_unmatched <- total - n_matched

cat(glue::glue("
Total SNPs: {total}
Matched original: {n_matched - n_flipped}
Matched flipped:  {n_flipped}
Unmatched:        {n_unmatched}
"))
```

# create a final scorefile and save as tsv
```{r}
final_scorefile <- scorefile %>%
  select(SNP = SNP_final, A1, BETA) %>%
  distinct(SNP, .keep_all = TRUE)

write_tsv(final_scorefile, glue("R:/Doktoranden/MarieFreudenberg/analysen/1kGp3_PRS/{trait}_hg38_scorefile_matched_and_flipped.txt"))
```

