---
title: "1000 Genomes FST Analysis"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook calculates Hudson's pairwise \(F_{ST}\) for the 1000 Genomes
Phase 3 dataset between individual populations and between superpopulations.


# Required input

## PLINK 2 genotype dataset

A PLINK 2 dataset with one shared file prefix:

```text
path/to/1kg_phase3.pgen
path/to/1kg_phase3.pvar
path/to/1kg_phase3.psam
```

The `.psam` file must contain these columns:

```text
#IID
Population
SuperPop
```

Each sample must have a population code such as `CEU`, `YRI`, or `CHB`, and a
superpopulation code such as `EUR`, `AFR`, or `EAS`.

The genotype dataset should already contain the variants intended for the FST
analysis, including any desired quality-control and MAF filters.

## Software

PLINK 2 is required:

<https://www.cog-genomics.org/plink/2.0/>

# 1. Load packages

```{r packages}
library(data.table)
library(dplyr)
library(tidyr)
library(pheatmap)
```

# 2. Set paths

```{r settings}
# Main project directory
path_prefix <- "path/to/project"

# PLINK 2 executable
plink2_path <- "path/to/plink2.exe"

# PLINK 2 genotype prefix without .pgen, .pvar, or .psam
genotype_prefix <- file.path(
  path_prefix,
  "path",
  "to",
  "1kg_phase3"
)

# Output directory
output_dir <- file.path(
  path_prefix,
  "1kGp3_FST"
)

dir.create(
  output_dir,
  recursive = TRUE,
  showWarnings = FALSE
)
```

# 3. Read sample information

```{r sample-information}
psam <- fread(
  paste0(genotype_prefix, ".psam")
)

if ("#IID" %in% names(psam)) {
  setnames(psam, "#IID", "IID")
}

required_columns <- c(
  "IID",
  "Population",
  "SuperPop"
)

missing_columns <- setdiff(
  required_columns,
  names(psam)
)

if (length(missing_columns) > 0) {
  stop(
    "The .psam file is missing these columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

psam <- psam %>%
  filter(
    !is.na(IID),
    !is.na(Population),
    !is.na(SuperPop)
  )

populations <- sort(
  unique(psam$Population)
)

superpopulations <- sort(
  unique(psam$SuperPop)
)

cat(
  "Samples:",
  nrow(psam),
  "\nPopulations:",
  length(populations),
  "\nSuperpopulations:",
  length(superpopulations),
  "\n"
)
```

# 4. Helper functions

```{r helper-functions}
run_plink <- function(arguments) {

  command <- paste(
    shQuote(plink2_path),
    arguments
  )

  cat("Running:\n", command, "\n\n")

  status <- system(command)

  if (status != 0) {
    stop(
      "PLINK 2 returned a non-zero exit status."
    )
  }
}

read_fst_summary <- function(path) {

  fst <- fread(path)

  names(fst) <- sub(
    "^#",
    "",
    names(fst)
  )

  names(fst) <- sub(
    "^X\\.",
    "",
    names(fst)
  )

  fst
}

fst_to_matrix <- function(
    fst_data,
    group_order
) {

  fst_matrix <- matrix(
    NA_real_,
    nrow = length(group_order),
    ncol = length(group_order),
    dimnames = list(
      group_order,
      group_order
    )
  )

  for (i in seq_len(nrow(fst_data))) {

    group1 <- fst_data$POP1[i]
    group2 <- fst_data$POP2[i]

    if (
      group1 %in% group_order &&
        group2 %in% group_order
    ) {
      fst_matrix[group1, group2] <-
        fst_data$HUDSON_FST[i]

      fst_matrix[group2, group1] <-
        fst_data$HUDSON_FST[i]
    }
  }

  diag(fst_matrix) <- NA_real_

  fst_matrix
}
```

# 5. Pairwise FST between populations

PLINK 2 reads the `Population` column directly from the `.psam` file.

```{r population-fst}
population_prefix <- file.path(
  output_dir,
  "populations"
)

run_plink(
  paste(
    "--pfile",
    shQuote(genotype_prefix),
    "--fst Population",
    "--out",
    shQuote(population_prefix)
  )
)

population_fst <- read_fst_summary(
  paste0(
    population_prefix,
    ".fst.summary"
  )
)

fwrite(
  population_fst,
  file.path(
    output_dir,
    "population_pairwise_fst.tsv"
  ),
  sep = "\t"
)
```

# 6. Population FST heatmap

The population order below groups populations by superpopulation.

```{r population-order}
population_order <- c(
  "ACB", "ASW", "ESN", "GWD", "LWK", "MSL", "YRI",
  "CLM", "MXL", "PEL", "PUR",
  "CDX", "CHB", "CHS", "JPT", "KHV",
  "CEU", "FIN", "GBR", "IBS", "TSI",
  "BEB", "GIH", "ITU", "PJL", "STU"
)

population_order <- population_order[
  population_order %in% populations
]

additional_populations <- setdiff(
  populations,
  population_order
)

population_order <- c(
  population_order,
  additional_populations
)

population_fst_matrix <- fst_to_matrix(
  population_fst,
  population_order
)
```

```{r population-heatmap}
pheatmap(
  population_fst_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
  number_format = "%.3f",
  main = "Pairwise Hudson FST between populations",
  fontsize_row = 8,
  fontsize_col = 8,
  filename = file.path(
    output_dir,
    "population_pairwise_fst_heatmap.png"
  ),
  width = 10,
  height = 9
)

pheatmap(
  population_fst_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
  number_format = "%.3f",
  main = "Pairwise Hudson FST between populations",
  fontsize_row = 8,
  fontsize_col = 8
)
```

# 7. Pairwise FST between superpopulations

PLINK 2 reads the `SuperPop` column directly from the `.psam` file.

```{r superpopulation-fst}
superpopulation_prefix <- file.path(
  output_dir,
  "superpopulations"
)

run_plink(
  paste(
    "--pfile",
    shQuote(genotype_prefix),
    "--fst SuperPop",
    "--out",
    shQuote(superpopulation_prefix)
  )
)

superpopulation_fst <- read_fst_summary(
  paste0(
    superpopulation_prefix,
    ".fst.summary"
  )
)

fwrite(
  superpopulation_fst,
  file.path(
    output_dir,
    "superpopulation_pairwise_fst.tsv"
  ),
  sep = "\t"
)
```

# 8. Superpopulation FST heatmap

```{r superpopulation-heatmap}
superpopulation_order <- c(
  "AFR",
  "AMR",
  "EAS",
  "EUR",
  "SAS"
)

superpopulation_order <-
  superpopulation_order[
    superpopulation_order %in%
      superpopulations
  ]

additional_superpopulations <- setdiff(
  superpopulations,
  superpopulation_order
)

superpopulation_order <- c(
  superpopulation_order,
  additional_superpopulations
)

superpopulation_fst_matrix <- fst_to_matrix(
  superpopulation_fst,
  superpopulation_order
)

pheatmap(
  superpopulation_fst_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
  number_format = "%.3f",
  main = "Pairwise Hudson FST between superpopulations",
  filename = file.path(
    output_dir,
    "superpopulation_pairwise_fst_heatmap.png"
  ),
  width = 7,
  height = 6
)

pheatmap(
  superpopulation_fst_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
  number_format = "%.3f",
  main = "Pairwise Hudson FST between superpopulations"
)
```