---
title: "Compare Population FST with PRS Differences"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Overview

This notebook compares pairwise genetic differentiation between populations,
measured with Hudson's \(F_{ST}\), with pairwise differences in mean polygenic
risk scores.

For every population pair, it:

1. calculates the mean PRS in each population;
2. calculates the absolute difference between the two population means;
3. joins this value to the corresponding pairwise \(F_{ST}\);
4. tests the correlation between \(F_{ST}\) and the PRS difference;
5. creates a scatter plot with a linear regression line.

# Required input

## 1. Population-level FST summary

Use the file created by `1kGp3_FST_combined.Rmd`:

```text
population_pairwise_fst.tsv
```

It must contain:

```text
POP1
POP2
HUDSON_FST
```

Each row represents one pair of 1000 Genomes populations.

## 2. Individual PRS file

A PLINK 2 `.sscore` file or tab-separated file containing one row per sample.

It must contain:

```text
IID
Population
SCORE1_SUM
```

- `IID` is the sample identifier.
- `Population` is the 1000 Genomes population code.
- `SCORE1_SUM` is the individual polygenic score.

If the population column is not already present in the score file, add it by
joining the score file to the 1000 Genomes `.psam` file before running this
notebook.

The score must be calculated for the same 1000 Genomes samples represented in
the FST analysis.

# 1. Load packages

```{r packages}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(glue)
```

# 2. Set paths

Change only the settings in this section.

```{r settings}
# Main project directory
path_prefix <- "path/to/project"

# Trait or score label used in output filenames and plot titles
trait <- "BMI"

# Individual PRS file
prs_file <- file.path(
  path_prefix,
  "1kGp3_PRS",
  paste0(
    trait,
    "_PRS_output.sscore"
  )
)

# Population pairwise FST file created by the FST notebook
fst_file <- file.path(
  path_prefix,
  "1kGp3_FST",
  "population_pairwise_fst.tsv"
)

# Output directory
output_dir <- file.path(
  path_prefix,
  "1kGp3_FST",
  "PRS_comparison"
)

dir.create(
  output_dir,
  recursive = TRUE,
  showWarnings = FALSE
)
```

# 3. Read and validate the PRS data

```{r read-prs}
prs <- read_tsv(
  prs_file,
  show_col_types = FALSE
)

required_prs_columns <- c(
  "IID",
  "Population",
  "SCORE1_SUM"
)

missing_prs_columns <- setdiff(
  required_prs_columns,
  names(prs)
)

if (length(missing_prs_columns) > 0) {
  stop(
    "The PRS file is missing these columns: ",
    paste(
      missing_prs_columns,
      collapse = ", "
    )
  )
}

prs <- prs %>%
  filter(
    !is.na(Population),
    !is.na(SCORE1_SUM)
  )
```

# 4. Calculate the mean PRS per population

```{r population-means}
prs_means <- prs %>%
  group_by(Population) %>%
  summarise(
    n_samples = n(),
    mean_prs = mean(
      SCORE1_SUM,
      na.rm = TRUE
    ),
    sd_prs = sd(
      SCORE1_SUM,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

write_tsv(
  prs_means,
  file.path(
    output_dir,
    paste0(
      trait,
      "_population_prs_summary.tsv"
    )
  )
)

prs_means
```

# 5. Calculate pairwise PRS differences

Only one row is retained for each unordered population pair.

```{r pairwise-prs-differences}
prs_differences <- prs_means %>%
  select(
    Population,
    mean_prs
  ) %>%
  rename(
    POP1 = Population,
    mean_prs_1 = mean_prs
  ) %>%
  crossing(
    prs_means %>%
      select(
        Population,
        mean_prs
      ) %>%
      rename(
        POP2 = Population,
        mean_prs_2 = mean_prs
      )
  ) %>%
  filter(
    POP1 < POP2
  ) %>%
  mutate(
    PRS_difference = abs(
      mean_prs_1 -
        mean_prs_2
    )
  )
```

# 6. Read and standardize the FST data

The population names are sorted within every pair so that the join is
independent of whether PLINK wrote a pair as `POP1–POP2` or `POP2–POP1`.

```{r read-fst}
fst <- read_tsv(
  fst_file,
  show_col_types = FALSE
)

names(fst) <- sub(
  "^#",
  "",
  names(fst)
)

names(fst) <- sub(
  "^X\\.",
  "",
  names(fst)
)

required_fst_columns <- c(
  "POP1",
  "POP2",
  "HUDSON_FST"
)

missing_fst_columns <- setdiff(
  required_fst_columns,
  names(fst)
)

if (length(missing_fst_columns) > 0) {
  stop(
    "The FST file is missing these columns: ",
    paste(
      missing_fst_columns,
      collapse = ", "
    )
  )
}

fst <- fst %>%
  transmute(
    POP1_original = POP1,
    POP2_original = POP2,
    POP1 = pmin(
      POP1_original,
      POP2_original
    ),
    POP2 = pmax(
      POP1_original,
      POP2_original
    ),
    HUDSON_FST = as.numeric(
      HUDSON_FST
    )
  ) %>%
  filter(
    !is.na(HUDSON_FST)
  )
```

# 7. Join FST and PRS differences

```{r join-data}
combined <- prs_differences %>%
  left_join(
    fst,
    by = c(
      "POP1",
      "POP2"
    )
  ) %>%
  filter(
    !is.na(HUDSON_FST),
    !is.na(PRS_difference)
  )

write_tsv(
  combined,
  file.path(
    output_dir,
    paste0(
      trait,
      "_population_FST_vs_PRS.tsv"
    )
  )
)

cat(
  "Population pairs with both FST and PRS data:",
  nrow(combined),
  "\n"
)
```

# 8. Correlation test

Pearson correlation is used to test the linear association. Spearman
correlation is also reported as a rank-based comparison.

```{r correlation-tests}
pearson_test <- cor.test(
  combined$HUDSON_FST,
  combined$PRS_difference,
  method = "pearson"
)

spearman_test <- cor.test(
  combined$HUDSON_FST,
  combined$PRS_difference,
  method = "spearman",
  exact = FALSE
)

correlation_results <- tibble(
  method = c(
    "Pearson",
    "Spearman"
  ),
  correlation = c(
    unname(
      pearson_test$estimate
    ),
    unname(
      spearman_test$estimate
    )
  ),
  p_value = c(
    pearson_test$p.value,
    spearman_test$p.value
  ),
  n_population_pairs = nrow(
    combined
  )
)

write_tsv(
  correlation_results,
  file.path(
    output_dir,
    paste0(
      trait,
      "_FST_PRS_correlation.tsv"
    )
  )
)

correlation_results
```

# 9. Plot FST against PRS differences

```{r plot}
fst_prs_plot <- ggplot(
  combined,
  aes(
    x = HUDSON_FST,
    y = PRS_difference
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  labs(
    title = glue(
      "{trait}: population PRS differences versus FST"
    ),
    x = "Pairwise Hudson FST",
    y = "Absolute difference in mean PRS"
  ) +
  theme_minimal()

fst_prs_plot

ggsave(
  filename = file.path(
    output_dir,
    paste0(
      trait,
      "_FST_vs_PRS.png"
    )
  ),
  plot = fst_prs_plot,
  width = 7,
  height = 5,
  dpi = 300
)
```