---
title: "Untitled"
output: html_document
date: "2025-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# load libraries and define base path 
```{r}
.libPaths("H:/Rlibs")
library(data.table)
library(readr)
library(dplyr)
library(glue)
library(pROC)
library(DescTools)
```

# Read summary statistics file
```{r}
trait <- "Diabetes"
study <- "Mahajan.NatGenet2018b.T2D.European"
trait_type <- "binary"   # "continuous" or "binary"

base_path <- glue("R:/Doktoranden/MarieFreudenberg/analysen/CS/{trait}/")
ss <- fread(file.path(glue("R:/Doktoranden/MarieFreudenberg/analysen/sumstats/{study}.tsv")))
head(ss)
```
# LD ref basepair positions are in hg19 
# If needed, add rsid to the sumstats ()
```{r}
ref <- fread("R:/Doktoranden/MarieFreudenberg/analysen/CS/ldblk_ukbb_eur/snpinfo_ukbb_hm3")

# Standardize columns
ss[, `:=`(
  chr38 = sub("^chr", "", as.character(chr)),          # original hg38 chr
  pos38 = as.integer(pos),                             # original hg38 pos
  chr19 = sub("^chr", "", as.character(chr_hg19)),     # lifted hg19 chr
  pos19 = as.integer(pos_hg19),                        # lifted hg19 pos  <- use this, not start_hg19
  a1    = toupper(a1),
  a0    = toupper(a0),
  beta  = as.numeric(beta),
  p     = as.numeric(p)
)]

ref[, `:=`(
  chr19 = sub("^chr", "", as.character(CHR)),
  pos19 = as.integer(BP),
  A1    = toupper(A1),
  A2    = toupper(A2)
)]

# Order-independent allele keys
ss[, `:=`(
  allele1 = pmin(a1, a0),
  allele2 = pmax(a1, a0)
)]

ref[, `:=`(
  allele1 = pmin(A1, A2),
  allele2 = pmax(A1, A2)
)]

# Keep only needed reference columns and remove duplicates
ref_map <- unique(
  ref[, .(chr19, pos19, allele1, allele2, SNP)]
)

# Merge rsID onto sumstats using hg19 coordinates
ss_annot <- merge(
  ss,
  ref_map,
  by = c("chr19", "pos19", "allele1", "allele2"),
  all.x = TRUE
)

# Keep only variants that got an rsID
ss <- ss_annot[!is.na(SNP) & SNP != ""]

cat("Number of sumstats variants with rsID:", nrow(ss), "\n")
head(ss)
```

# sumstats need to have: SNP, A1, A2, BETA, SE (or P instead of SE)
```{r}
prscs_ss <- ss[
  !is.na(rsid) & !is.na(beta) & !is.na(p),
  .(
    SNP  = rsid,
    A1   = a1,
    A2   = a0,
    BETA = beta,
    P    = p
  )
]

fwrite(
  prscs_ss,
  file = file.path(base_path, glue("{study}_PRScs.txt")),
  sep = "\t",
  quote = FALSE,
  na = "NA"
)
```

# Add rsid to LIFE-Adult bim files(use rsid from the sumstats - keep only snps from sumstats)
```{r}
bim <- fread("R:/Doktoranden/MarieFreudenberg/analysen/CS/LIFE_bed.bim")
setnames(bim, c("CHR", "SNP_old", "CM", "POS", "A1", "A2"))

# make join columns consistent immediately
bim[, `:=`(
  CHR = as.character(CHR),
  POS = as.integer(POS),
  A1  = toupper(A1),
  A2  = toupper(A2)
)]

bim[, `:=`(
  allele1 = pmin(A1, A2),
  allele2 = pmax(A1, A2)
)]

ss_map <- unique(
  ss[
    !is.na(rsid) & !is.na(chr) & !is.na(pos),
    .(
      CHR = as.character(chr),
      POS = as.integer(pos),
      allele1 = as.character(a1),
      allele2 = as.character(a0),
      rsid = rsid
    )
  ]
)

map <- merge(
  bim[, .(CHR, POS, allele1, allele2, SNP_old)],
  ss_map,
  by = c("CHR", "POS", "allele1", "allele2")
)

renamer <- unique(map[, .(oldID = SNP_old, newID = rsid)])

fwrite(
  renamer,
  file = file.path(base_path, glue("{study}_rename_to_rsid.txt")),
  sep = "\t",
  col.names = FALSE,
  quote = FALSE
)

keep_snps_new <- unique(map$rsid)

fwrite(
  data.table(SNP = keep_snps_new),
  file = file.path(base_path, glue("{study}_rsid.txt")),
  sep = "\t",
  col.names = FALSE,
  quote = FALSE
)
```

```{r}
extract_file <- glue("{base_path}/{study}_rsid.txt")
rename_file  <- glue("{base_path}/{study}_rename_to_rsid.txt")
out_prefix   <- glue("{base_path}/{study}_LIFE_rsid")

cmd <- paste(
  shQuote("C:/Users/mar5862fre/Downloads/plink2_win64_20251026/plink2.exe"),
  "--bfile", shQuote("R:/Doktoranden/MarieFreudenberg/analysen/CS/LIFE_bed"),
  "--extract", shQuote(extract_file),
  "--update-name", shQuote(rename_file), "2", "1",
  "--make-bed",
  "--out", shQuote(out_prefix)
)

cat("Running:\n", cmd, "\n")
system(cmd)
```

# run this command in python in PRScs-master directory 
# python PRScs.py --ref_dir=PATH_TO_REFERENCE --bim_prefix=VALIDATION_BIM_PREFIX --sst_file=SUM_STATS_FILE --n_gwas=GWAS_SAMPLE_SIZE --out_dir=OUTPUT_DIR [--a=PARAM_A --b=PARAM_B --phi=PARAM_PHI --n_iter=MCMC_ITERATIONS --n_burnin=MCMC_BURNIN --thin=MCMC_THINNING_FACTOR --chrom=CHROM --beta_std=BETA_STD --write_psi=WRITE_PSI --write_pst=WRITE_POSTERIOR_SAMPLES --seed=SEED]

# print full command (manually adjust GWAS sample size)
```{r}
bim_prefix <- out_prefix
sst_file   <- glue("{base_path}/{study}_PRScs.txt")
out_dir    <- glue("{base_path}/{study}_prscs")

cmd <- paste(
  "python PRScs.py",
  paste0("--ref_dir=", shQuote("R:/Doktoranden/MarieFreudenberg/analysen/CS/ldblk_1kg_eur")),
  paste0("--bim_prefix=", shQuote(bim_prefix)),
  paste0("--sst_file=", shQuote(sst_file)),
  "--n_gwas=900000",
  paste0("--out_dir=", shQuote(out_dir)),
  "--seed=1234"
)

cat(cmd, "\n")
```

# concatenate the 22 chr files into one big file and loose the columns that are not neccesary to score 
```{r}
# Find all PRS-CS effect files and stack them 
files <- list.files(
  path = base_path,
  pattern = paste0("^", study, "_prscs_pst_eff"),
  full.names = TRUE
)

length(files)        # should be 22
files[1:5]

allchr <- rbindlist(lapply(files, fread))

# Rename columns for clarity
setnames(allchr, c("CHR", "SNP", "BP", "A1", "A2", "BETA"))

# Check overlap with BIM
bim <- fread(file.path(glue(base_path, "{study}_LIFE_rsid.bim")))
setnames(bim, c("CHR", "SNP", "CM", "BP", "A1", "A2"))

cat("Score SNPs:", length(unique(allchr$SNP)), "\n")
cat("BIM SNPs:  ", length(unique(bim$SNP)), "\n")
cat("Overlap:   ", length(intersect(allchr$SNP, bim$SNP)), "\n")

# Create simple 3-column score file
score <- allchr[, .(SNP, A1, BETA)]

fwrite(
  score,
  file = file.path(glue(base_path, "{study}_prscs_scorefile.txt")),
  sep = "\t"
)
```

# score LIFE data
```{r}
score_file <- file.path(base_path, glue("{study}_prscs_scorefile.txt"))
out_file   <- file.path(base_path, glue("{study}_individual_scores"))

cmd <- paste(
  shQuote("C:/Users/mar5862fre/Downloads/plink2_win64_20251026/plink2.exe"),
  "--bfile", shQuote(out_prefix),
  "--score", shQuote(score_file),
  "1 2 3 header-read cols=+scoresums",
  "--out", shQuote(out_file)
)

cat(cmd, "\n")
system(cmd)
```

# Create file with phenotype, metadata and prs scores 
# load files 
```{r}
data <- fread(glue("R:/Doktoranden/MarieFreudenberg/analysen/Phenofiles/{trait}_covars.tsv"))
prs <- read_table(glue(base_path, "{study}_individual_scores.sscore"))
```
# Add PRS to covars data
```{r}
data <- data %>%
  left_join(prs, by = "IID")
data <- data[complete.cases(data), ]
```

# Run linear/logistic regression and check how much variance is explained by PRS
```{r}
outcome_var <- "Phenotype"
prs_var     <- "BETA_SUM"
covars      <- c("Age", "Sex", paste0("PC", 1:10))

analysis_vars <- c(outcome_var, prs_var, covars)

analysis_data <- as.data.frame(data[, ..analysis_vars])
analysis_data <- analysis_data[complete.cases(analysis_data), ]

analysis_data$PRS_SD <- as.numeric(scale(analysis_data[[prs_var]]))

base_formula <- as.formula(
  paste(outcome_var, "~", paste(covars, collapse = " + "))
)

full_formula <- as.formula(
  paste(outcome_var, "~", paste(c("PRS_SD", covars), collapse = " + "))
)

trait_type <- tolower(trait_type)

if (trait_type == "continuous") {

  # Linear regression
  model_reduced <- lm(base_formula, data = analysis_data)
  model_full    <- lm(full_formula, data = analysis_data)

  r2_base <- summary(model_reduced)$r.squared
  r2_full <- summary(model_full)$r.squared
  r2_prs  <- r2_full - r2_base

  prs_row <- coef(summary(model_full))["PRS_SD", ]
  prs_ci  <- confint(model_full, "PRS_SD")

  var_explained_txt <- glue("
Base model R²:                {round(r2_base, 4)}
Full model R²:                {round(r2_full, 4)}
Incremental R² (PRS):         {round(r2_prs, 4)}

PRS effect per SD:
  Beta:                       {round(unname(prs_row['Estimate']), 4)}
  95% CI:                     [{round(prs_ci[1], 4)}, {round(prs_ci[2], 4)}]
  P-value:                    {signif(unname(prs_row['Pr(>|t|)']), 4)}
")

  cat(var_explained_txt, "\n")
  print(summary(model_full))

  results <- list(
    trait_type   = trait_type,
    n            = nrow(analysis_data),
    base_model   = model_reduced,
    full_model   = model_full,
    r2_base      = r2_base,
    r2_full      = r2_full,
    r2_prs       = r2_prs,
    beta_per_sd  = unname(prs_row["Estimate"]),
    beta_ci_low  = prs_ci[1],
    beta_ci_high = prs_ci[2],
    p_value      = unname(prs_row["Pr(>|t|)"])
  )

} else if (trait_type == "binary") {

  # Make sure the outcome is binary
  y <- analysis_data[[outcome_var]]

  if (is.factor(y) || is.character(y)) {
    analysis_data[[outcome_var]] <- droplevels(as.factor(y))
    if (nlevels(analysis_data[[outcome_var]]) != 2) {
      stop("For trait_type = 'binary', the phenotype must have exactly 2 levels.")
    }
  } else {
    uniq_y <- sort(unique(y))
    if (length(uniq_y) != 2) {
      stop("For trait_type = 'binary', the phenotype must have exactly 2 values.")
    }
    # Recode numeric/logical outcomes to 0/1 using the higher value as the event
    analysis_data[[outcome_var]] <- ifelse(y == uniq_y[2], 1, 0)
  }

  # Logistic regression
  model_reduced <- glm(base_formula, data = analysis_data, family = binomial())
  model_full    <- glm(full_formula, data = analysis_data, family = binomial())

  # Predicted probabilities
  prob_base <- predict(model_reduced, type = "response")
  prob_full <- predict(model_full, type = "response")

  # AUC
  roc_base <- pROC::roc(response = analysis_data[[outcome_var]], predictor = prob_base, quiet = TRUE)
  roc_full <- pROC::roc(response = analysis_data[[outcome_var]], predictor = prob_full, quiet = TRUE)

  auc_base <- as.numeric(pROC::auc(roc_base))
  auc_full <- as.numeric(pROC::auc(roc_full))
  auc_diff <- auc_full - auc_base

  # Nagelkerke R²
  nag_base <- unname(DescTools::PseudoR2(model_reduced, which = "Nagelkerke"))
  nag_full <- unname(DescTools::PseudoR2(model_full, which = "Nagelkerke"))
  nag_diff <- nag_full - nag_base

  # PRS effect per SD
  prs_row   <- coef(summary(model_full))["PRS_SD", ]
  log_or    <- unname(prs_row["Estimate"])
  se_log_or <- unname(prs_row["Std. Error"])
  p_value   <- unname(prs_row["Pr(>|z|)"])

  or_per_sd <- exp(log_or)
  or_ci     <- exp(c(log_or - 1.96 * se_log_or, log_or + 1.96 * se_log_or))
  odds_increase_pct <- (or_per_sd - 1) * 100

  logistic_txt <- glue("
Base model AUC:               {round(auc_base, 4)}
Full model AUC:               {round(auc_full, 4)}
Delta AUC (PRS):              {round(auc_diff, 4)}

Base model Nagelkerke R²:     {round(nag_base, 4)}
Full model Nagelkerke R²:     {round(nag_full, 4)}
Delta Nagelkerke R² (PRS):    {round(nag_diff, 4)}

PRS effect per SD:
  Log-odds beta:              {round(log_or, 4)}
  OR per SD:                  {round(or_per_sd, 4)}
  95% CI for OR:              [{round(or_ci[1], 4)}, {round(or_ci[2], 4)}]
  % increase in odds / SD:    {round(odds_increase_pct, 2)}%
  P-value:                    {signif(p_value, 4)}
")

  cat(logistic_txt, "\n")
  print(summary(model_full))

  results <- list(
    trait_type         = trait_type,
    n                  = nrow(analysis_data),
    base_model         = model_reduced,
    full_model         = model_full,
    auc_base           = auc_base,
    auc_full           = auc_full,
    auc_diff           = auc_diff,
    nagelkerke_base    = nag_base,
    nagelkerke_full    = nag_full,
    nagelkerke_diff    = nag_diff,
    log_or_per_sd      = log_or,
    or_per_sd          = or_per_sd,
    or_ci_low          = or_ci[1],
    or_ci_high         = or_ci[2],
    odds_increase_pct  = odds_increase_pct,
    p_value            = p_value
  )

}
```


