---
title: "PRS-CS"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This notebook prepares GWAS summary statistics and target genotype data for
PRS-CS, runs PRS-CS, combines the chromosome-specific effect files, and
calculates individual polygenic scores with PLINK 2.

PRS-CS is available from:

<https://github.com/getian107/PRScs>

## Required input

- GWAS summary statistics containing chromosome, position, alleles, effect
  estimate, p-value, and hg19 coordinates.
- A target genotype dataset in PLINK BED format:
  `.bed`, `.bim`, and `.fam`.
- A phenotype and covariate file containing:
  `IID`, `Phenotype`, `Age`, `Sex`, and `PC1` to `PC10`.
- The PRS-CS source code downloaded from the official repository.
- A PRS-CS LD reference panel downloaded through the official repository.
- Python with the required PRS-CS dependencies.
- PLINK 2.

# Load libraries

```{r}
library(data.table)
library(readr)
library(dplyr)
library(glue)
library(pROC)
library(DescTools)
```

# Set paths and analysis options

Change only the settings in this section.

```{r}
# Main project directory
path_prefix <- "path/to/project"

trait <- "Diabetes"
study <- "Mahajan.NatGenet2018b.T2D.European"
trait_type <- "binary"  # "continuous" or "binary"

# GWAS sample size used by PRS-CS
n_gwas <- 900000

# Executables and software
plink2_path <- "path/to/plink2.exe"
python_path <- "python"

# PRS-CS repository directory
prscs_dir <- file.path(
  path_prefix,
  "PRScs"
)

# PRS-CS LD reference directory
ld_reference_dir <- file.path(
  path_prefix,
  "PRS_CS",
  "ldblk_1kg_eur"
)

# SNP information file from the selected reference panel
reference_snpinfo_file <- file.path(
  ld_reference_dir,
  "snpinfo_1kg_hm3"
)

# Input files
sumstats_file <- file.path(
  path_prefix,
  "sumstats",
  paste0(study, ".tsv")
)

genotype_prefix <- file.path(
  path_prefix,
  "PRS_CS",
  "LIFE_bed"
)

phenotype_file <- file.path(
  path_prefix,
  "Phenofiles",
  paste0(trait, "_covars.tsv")
)

# Analysis output directory
base_path <- file.path(
  path_prefix,
  "PRS_CS",
  trait
)

dir.create(
  base_path,
  recursive = TRUE,
  showWarnings = FALSE
)
```

# Read summary statistics

```{r}
ss <- fread(sumstats_file)

head(ss)
```

# If needed, add rsIDs using the PRS-CS reference panel

The PRS-CS LD reference positions are in hg19. The summary statistics are
therefore matched to the reference using hg19 chromosome, position, and
alleles.

```{r}
ref <- fread(reference_snpinfo_file)

ss[, `:=`(
  chr38 = sub("^chr", "", as.character(chr)),
  pos38 = as.integer(pos),
  chr19 = sub("^chr", "", as.character(chr_hg19)),
  pos19 = as.integer(pos_hg19),
  a1 = toupper(a1),
  a0 = toupper(a0),
  beta = as.numeric(beta),
  p = as.numeric(p)
)]

ref[, `:=`(
  chr19 = sub("^chr", "", as.character(CHR)),
  pos19 = as.integer(BP),
  A1 = toupper(A1),
  A2 = toupper(A2)
)]

# Create order-independent allele keys
ss[, `:=`(
  allele1 = pmin(a1, a0),
  allele2 = pmax(a1, a0)
)]

ref[, `:=`(
  allele1 = pmin(A1, A2),
  allele2 = pmax(A1, A2)
)]

# Keep the required reference columns
ref_map <- unique(
  ref[, .(
    chr19,
    pos19,
    allele1,
    allele2,
    rsid = SNP
  )]
)

# Add rsIDs to the summary statistics
ss <- merge(
  ss,
  ref_map,
  by = c(
    "chr19",
    "pos19",
    "allele1",
    "allele2"
  ),
  all.x = TRUE
)

# Retain variants that received an rsID
ss <- ss[
  !is.na(rsid) &
    rsid != ""
]

cat(
  "Number of summary-statistic variants with rsID:",
  nrow(ss),
  "\n"
)

head(ss)
```

# Create the PRS-CS summary-statistics file

PRS-CS accepts `BETA` and `P` together with the SNP and allele columns.

```{r}
prscs_ss <- ss[
  !is.na(rsid) &
    !is.na(beta) &
    !is.na(p),
  .(
    SNP = rsid,
    A1 = a1,
    A2 = a0,
    BETA = beta,
    P = p
  )
]

prscs_sumstats_file <- file.path(
  base_path,
  paste0(study, "_PRScs.txt")
)

fwrite(
  prscs_ss,
  file = prscs_sumstats_file,
  sep = "\t",
  quote = FALSE,
  na = "NA"
)
```

# Add rsIDs to the target BIM file

The target genotype data are restricted to variants present in the summary
statistics and renamed to the corresponding rsIDs.

```{r}
bim <- fread(
  paste0(genotype_prefix, ".bim")
)

setnames(
  bim,
  c(
    "CHR",
    "SNP_old",
    "CM",
    "POS",
    "A1",
    "A2"
  )
)

bim[, `:=`(
  CHR = as.character(CHR),
  POS = as.integer(POS),
  A1 = toupper(A1),
  A2 = toupper(A2)
)]

bim[, `:=`(
  allele1 = pmin(A1, A2),
  allele2 = pmax(A1, A2)
)]

ss_map <- unique(
  ss[
    !is.na(rsid) &
      !is.na(chr) &
      !is.na(pos),
    .(
      CHR = as.character(chr),
      POS = as.integer(pos),
      allele1 = pmin(a1, a0),
      allele2 = pmax(a1, a0),
      rsid = rsid
    )
  ]
)

map <- merge(
  bim[, .(
    CHR,
    POS,
    allele1,
    allele2,
    SNP_old
  )],
  ss_map,
  by = c(
    "CHR",
    "POS",
    "allele1",
    "allele2"
  )
)

renamer <- unique(
  map[, .(
    oldID = SNP_old,
    newID = rsid
  )]
)

rename_file <- file.path(
  base_path,
  paste0(study, "_rename_to_rsid.txt")
)

fwrite(
  renamer,
  file = rename_file,
  sep = "\t",
  col.names = FALSE,
  quote = FALSE
)

keep_snps_new <- unique(map$rsid)

extract_file <- file.path(
  base_path,
  paste0(study, "_rsid.txt")
)

fwrite(
  data.table(SNP = keep_snps_new),
  file = extract_file,
  sep = "\t",
  col.names = FALSE,
  quote = FALSE
)
```

# Create the target genotype files for PRS-CS

```{r}
out_prefix <- file.path(
  base_path,
  paste0(study, "_LIFE_rsid")
)

cmd <- paste(
  shQuote(plink2_path),
  "--bfile",
  shQuote(genotype_prefix),
  "--extract",
  shQuote(extract_file),
  "--update-name",
  shQuote(rename_file),
  "2",
  "1",
  "--make-bed",
  "--out",
  shQuote(out_prefix)
)

cat("Running:\n", cmd, "\n")

system(cmd)
```

# Run PRS-CS

The command is printed here and can be run from the PRS-CS repository
directory.

```{r}
prscs_output_prefix <- file.path(
  base_path,
  paste0(study, "_prscs")
)

prscs_script <- file.path(
  prscs_dir,
  "PRScs.py"
)

cmd <- paste(
  shQuote(python_path),
  shQuote(prscs_script),
  paste0(
    "--ref_dir=",
    shQuote(ld_reference_dir)
  ),
  paste0(
    "--bim_prefix=",
    shQuote(out_prefix)
  ),
  paste0(
    "--sst_file=",
    shQuote(prscs_sumstats_file)
  ),
  paste0(
    "--n_gwas=",
    n_gwas
  ),
  paste0(
    "--out_dir=",
    shQuote(prscs_output_prefix)
  ),
  "--seed=1234"
)

cat(cmd, "\n")
```

# Combine the chromosome-specific PRS-CS output files

Run this section after the PRS-CS command has finished.

```{r}
files <- list.files(
  path = base_path,
  pattern = paste0(
    "^",
    study,
    "_prscs_pst_eff"
  ),
  full.names = TRUE
)

length(files)

allchr <- rbindlist(
  lapply(
    files,
    fread
  )
)

setnames(
  allchr,
  c(
    "CHR",
    "SNP",
    "BP",
    "A1",
    "A2",
    "BETA"
  )
)
```

# Check overlap with the target BIM file

```{r}
bim <- fread(
  paste0(out_prefix, ".bim")
)

setnames(
  bim,
  c(
    "CHR",
    "SNP",
    "CM",
    "BP",
    "A1",
    "A2"
  )
)

cat(
  "Score SNPs:",
  length(unique(allchr$SNP)),
  "\n"
)

cat(
  "BIM SNPs:",
  length(unique(bim$SNP)),
  "\n"
)

cat(
  "Overlap:",
  length(
    intersect(
      allchr$SNP,
      bim$SNP
    )
  ),
  "\n"
)
```

# Create the PRS-CS score file

```{r}
score <- allchr[
  ,
  .(
    SNP,
    A1,
    BETA
  )
]

score_file <- file.path(
  base_path,
  paste0(
    study,
    "_prscs_scorefile.txt"
  )
)

fwrite(
  score,
  file = score_file,
  sep = "\t"
)
```

# Calculate individual scores with PLINK 2

```{r}
individual_score_prefix <- file.path(
  base_path,
  paste0(
    study,
    "_individual_scores"
  )
)

cmd <- paste(
  shQuote(plink2_path),
  "--bfile",
  shQuote(out_prefix),
  "--score",
  shQuote(score_file),
  "1 2 3 header-read cols=+scoresums",
  "--out",
  shQuote(individual_score_prefix)
)

cat(cmd, "\n")

system(cmd)
```

# Merge the PRS with the phenotype file

```{r}
data <- fread(
  phenotype_file
)

prs <- read_table(
  paste0(
    individual_score_prefix,
    ".sscore"
  )
)

data <- data %>%
  left_join(
    prs,
    by = "IID"
  )

data <- data[
  complete.cases(data),
]
```
