---
title: "snpboost"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This notebook fits a polygenic score using `snpboost` and individual-level
genotype and phenotype data.

The phenotype file is assumed to have already been prepared, as in the other
analysis scripts.

## Required input

The phenotype file must contain at least:

```text
IID
Phenotype
Age
Sex
```

The target genotype dataset must be available in PLINK 2 format:

```text
genotype_prefix.pgen
genotype_prefix.pvar
genotype_prefix.psam
```

The following snpboost source files are also required and can be downloaded 
from https://github.com/hklinkhammer/snpboost:

```text
snpboost.R
functions_snpboost.R
functions_snpnet.R
lm_boost.R
```

PLINK 2 and `zstdcat` must be installed locally.

# Load libraries

```{r}
library(dplyr)
library(readr)
library(glue)
```

# Set paths and analysis options

Change only the settings in this section.

```{r}
# Main project directory
path_prefix <- "path/to/project"

# Trait name
trait <- "BMI"

# Input phenotype and covariate file
phenotype_file <- file.path(
  path_prefix,
  "Phenofiles",
  paste0(trait, "_covars.tsv")
)

# PLINK 2 genotype prefix without a file extension
genotype_prefix <- file.path(
  path_prefix,
  "path",
  "to",
  "genotype_data"
)

# Directory containing the snpboost source files
snpboost_dir <- file.path(
  path_prefix,
  "snpboost"
)

# External executables
plink2_path <- "path/to/plink2.exe"
zstdcat_path <- "path/to/zstdcat.exe"

# Output directory
results_dir <- file.path(
  path_prefix,
  "snpboost",
  "snpboost_results",
  trait
)
```

# Read the phenotype file

```{r}
pheno <- read_tsv(
  phenotype_file,
  show_col_types = FALSE
) %>%
  select(
    IID,
    Phenotype,
    Age,
    Sex
  ) %>%
  filter(
    !is.na(IID),
    !is.na(Phenotype),
    !is.na(Age),
    !is.na(Sex)
  )
```

# Create training, validation, and test sets

The data are divided into:

- 60% training data;
- 20% validation data;
- 20% test data.

```{r}
set.seed(42)

n <- nrow(pheno)
idx <- sample(seq_len(n))

train_end <- round(0.6 * n)
validation_end <- round(0.8 * n)

train_idx <- idx[1:train_end]
val_idx <- idx[(train_end + 1):validation_end]
test_idx <- idx[(validation_end + 1):n]

pheno$split <- NA_character_
pheno$split[train_idx] <- "train"
pheno$split[val_idx] <- "val"
pheno$split[test_idx] <- "test"
```

# Save the snpboost phenotype input file

```{r}
snpboost_pheno_file <- file.path(
  snpboost_dir,
  paste0(trait, "_input.tsv")
)

write_tsv(
  pheno,
  snpboost_pheno_file
)
```

# Load the snpboost functions

```{r}
source(
  file.path(
    snpboost_dir,
    "snpboost.R"
  )
)

source(
  file.path(
    snpboost_dir,
    "functions_snpboost.R"
  )
)

source(
  file.path(
    snpboost_dir,
    "functions_snpnet.R"
  )
)

source(
  file.path(
    snpboost_dir,
    "lm_boost.R"
  )
)
```

# Fit the snpboost model

```{r}
prs_model <- snpboost(
  genotype.pfile = genotype_prefix,
  phenotype.file = snpboost_pheno_file,
  phenotype = "Phenotype",
  covariates = c(
    "Age",
    "Sex"
  ),
  split.col = "split",
  family = "gaussian",
  metric = "MSEP",
  p_batch = 1000,
  m_batch = 1000,
  b_stop = 1,
  sl = 0.01,
  configs = list(
    plink2.path = plink2_path,
    zstdcat.path = zstdcat_path,
    nCores = parallel::detectCores() - 1,
    results.dir = results_dir,
    meta.dir = "meta",
    save.dir = "results",
    save = TRUE,
    save.computeProduct = TRUE
  )
)
```
